### 2023-07-02 Article.tex
First upload
This is the first upload

%% Author_tex.tex
%% V1.0
%% 2016/09/02
%% developed by Techset
%%
%% This file describes the coding for cta-author.cls

\documentclass{cta-author}%%%%where cta-author is the template name

\newtheorem{theorem}{Theorem}{}
\newtheorem{corollary}{Corollary}{}
\newtheorem{remark}{Remark}{}

%%%%%%Begin MyPackages%%
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{framed}
\usepackage{multirow}
\usepackage{color}
\usepackage{tabulary}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{url}
\usepackage{soul}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{caption}
\usepackage[title]{appendix}
\usepackage{ulem}
\usepackage{placeins}
\usepackage{balance}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
%%%%%%end MyPackages%%%%

\begin{document}

\title{How does an Experimental Software Engineering Lab Work? An Ethnographic Study}

\author{\au{Efraín R. Fonseca C.$^{1\corr}$}, \au{Oscar Dieste$^{2}$}, \au{Natalia Juristo$^{2}$}}

\address{\add{1}{Departamento de Ciencias de la Computación, Universidad de las Fuerzas Armadas - ESPE, Av. General Rumiñahui s/n y Ambato, Quito, Ecuador}
\add{2}{Departamento de Lenguajes y Sistemas Informáticos e Ingeniería de Software, Universidad Politécnica de Madrid, Calle de los Ciruelos, 28660, Madrid, España}
%\add{3}{Third Department, Third University, Address, Country Name}
%\add{4}{Current affiliation: Fourth Department, Fourth University, Address, Country Name}
\email{erfonseca@espe.edu.ec}}

\input{tex/abstract}

\maketitle

%%%%%%% Begin Content %%%%%%%
\input{tex/introduction}
\input{tex/research-method}
\input{tex/research-execution}
\input{tex/results}
\input{tex/discussion}
\input{tex/threats-to-validity}
\input{tex/conclusion}
\input{tex/data-availability-statements}
%%%%%%% End Content %%%%%%%%%

\begin{comment}

\section{Insert A head here}
This demo file is intended to serve as a ``starter file''
for IMAIAI journal papers produced under \LaTeX\ using
imaiai.cls v1.5e.

\subsection{Insert B head here}
Subsection text here.

\subsubsection{Insert C head here}
Subsubsection text here.

\section{Equations}

Sample equations.

%%% Numbered equation
\begin{align}\label{1.1}
\begin{split}
\frac{\partial u(t,x)}{\partial t} &= Au(t,x) \left(1-\frac{u(t,x)}{K}\right)-B\frac{u(t-\tau,x) w(t,x)}{1+Eu(t-\tau,x)},\\
\frac{\partial w(t,x)}{\partial t} &=\delta \frac{\partial^2w(t,x)}{\partial x^2}-Cw(t,x)+D\frac{u(t-\tau,x)w(t,x)}{1+Eu(t-\tau,x)},
\end{split}
\end{align}

\begin{align}\label{1.2}
\begin{split}
\frac{dU}{dt} &=\alpha U(t)(\gamma -U(t))-\frac{U(t-\tau)W(t)}{1+U(t-\tau)},\\
\frac{dW}{dt} &=-W(t)+\beta\frac{U(t-\tau)W(t)}{1+U(t-\tau)}.
\end{split}
\end{align}

%%%% Unnumbered equation
\[
\frac{\partial(F_1,F_2)}{\partial(c,\omega)}_{(c_0,\omega_0)} = \left|
\begin{array}{ll}
\frac{\partial F_1}{\partial c} &\frac{\partial F_1}{\partial \omega} \\\noalign{\vskip3pt}
\frac{\partial F_2}{\partial c}&\frac{\partial F_2}{\partial \omega}
\end{array}\right|_{(c_0,\omega_0)}
\]

\section{Enunciations}
%%%% Most of the enunciations like theorem, lemma, corollary, proposition, defintion,
%%%% condition, example, conjecture etc. are defined in the class file.

%%%% If the author wants to add or modify the enunciation style
%%%% they can define in the preamble as shown below.

%%%% \newtheoremstyle{theorem}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%% \theoremstyle{theorem}
%%%% \newtheorem{theorem}{\sc Theorem}[section]

%%%%\newtheoremstyle{corollary}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%%\theoremstyle{corollary}
%%%%\newtheorem{corollary}{\sc Corollary}[section]

%%%%\newtheoremstyle{definition}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%%\theoremstyle{definition}
%%%%\newtheorem{definition}[theorem]{\sc Definition}
%%%%
%%%%\newtheorem{exercise}[theorem]{Exercise}

\begin{theorem}\label{T0.1}
Assume that $\alpha>0, \gamma>1, \beta>\frac{\gamma+1}{\gamma-1}$.
Then there exists a small $\tau_1>0$, such that for $\tau\in
[0,\tau_1)$, if $c$ crosses $c(\tau)$ from the direction of
to  a small amplitude periodic traveling wave solution of
(2.1), and the period of $(\check{u}^p(s),\check{w}^p(s))$ is
\[
\check{T}(c)=c\cdot \left[\frac{2\pi}{\omega(\tau)}+O(c-c(\tau))\right].
\]
\end{theorem}

\section{Figures \& Tables}

The output for figure is:

\begin{figure}[!h]
%\centering\includegraphics[width=2.5in]{figurename.eps}
%%%call your figure name in the place "figurename.eps"
\caption{Insert figure caption here}
\label{fig_sim}
\end{figure}

 An example of a double column floating figure using two subfigures.
 (The subfig.sty package must be loaded for this to work.)
 The subfigure \verb+\label+ commands are set within each subfloat command, the
 \verb+\label+ for the overall figure must come after \verb+\caption+.
 \verb+\hfil+ must be used as a separator to get equal spacing.
 The subfigure.sty package works much the same way, except \verb+\subfigure+ is
 used instead of \verb+\subfloat+.

%\begin{figure*}[!h]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{figurename.eps}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{figurename.eps}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}

\vskip2pc

\noindent The output for table is:

\begin{table}[!h]
\caption{An Example of a Table}%%%Table caption goes here
\label{table_example}
\centering
\begin{tabular}{|c||c|}%%%The number of columns has to be defined here
\hline
One & Two\\ %%%% Table body
\hline
Three & Four\\%%%% Table body
\hline
\end{tabular}
\end{table}%%%End of the table

\section{Conclusion}
The conclusion text goes here.
\end{comment}

\section*{Acknowledgment}

Insert the Acknowledgment text here.

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/

\bibliographystyle{imaiai}
\bibliography{bibliography}
%
% once the .bbl file has been generated then place the text in your article.

% To get the numbered reference style the author should use [numbib]
%as an option in the document class.  For example: \documentclass[numbib]{imaiai}
\begin{comment}

\begin{thebibliography}{9}
\bibitem{1}
Vicsek T., Czir\'{o}k A., Ben-Jacob E., Cohen I.: `Novel type of phase transition in a system of self-driven particles', \textit{Phys. Rev. Lett.}, 1995, \textbf{75}, pp.~1226--1229

\bibitem{2}
Jadbabaie A., Lin J., Morse A.S.: `Coordination of groups of
mobile autonomous agents using nearest neighbor rules', \textit{IEEE
Trans. Autom. Control}, 2003, \textbf{48}, pp.~988--1001

\bibitem{3}
Olfati-Saber R., Murray R.M.: `Consensus problems in networks
of agents with switching topology and time-delays', \textit{IEEE Trans.
Autom. Control}, 2004, \textbf{49}, pp.~1520--1533
\end{thebibliography}
\end{comment}

\end{document}

### 2023-07-02 tex/abstract.tex
First upload
This is the first upload

\begin{abstract}
\textbf{Context:} Experimental replication is complex in Software Engineering. Several strategies, such as replication packages and families of experiments, have been applied with limited success. We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, etc., or intrinsic reasons, i.e., the researchers' needs are not fulfilled by the proposed replication procedures.
\textbf{Objective:} To find out how ESE researchers conduct experiments in practice.
\textbf{Method:} We carried out an ethnographical study with an experimental research group for two years instead of working halfway with several research groups over short periods.
\textbf{Results:} We have created conceptual and process models representing how experimentation, replication and synthesis are conducted in the target research group. These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels. The experimental process carried out in the group differs from textbooks in terms of (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different sub-areas or families of experiments have about the experimental process.
\textbf{Conclusions:} The differences between the actual lab processes vs the idealized versions portrayed in textbooks probably harm knowledge transfer among labs and make replication harder. However, our observations are limited to one research group. We plan to conduct further research with a representative sample of the research groups, e.g., using a survey, to explore whether our findings have general applicability.[CHECK NO MORE THAN 200 WORDS]
\end{abstract}

### 2023-07-02 tex/conclusion.tex
First upload
This is the first upload

\section{Conclusions}\label{sec-conclusions}
Experimental researchers know that experimentation, as described in textbooks and articles, is only an approximation to the daily work in the labs and the field. This study shows the deviations found in a concrete experimental research group. Such divergences do not seem arbitrary but necessary from a systemic viewpoint, e.g., each people have distinct abilities and perform different tasks in a research project.

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group in which we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the labs and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k} \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests at a relatively superficial level, e.g., when researchers from different experimental families discuss their mental models.

Furthermore, anecdotic evidence, e.g., conversations with colleagues, suggests that this "lab mythology" can also be somewhat present in SE. We all know that the experiment reports do not reflect the experimental process precisely. It does not mean experimenters cheat. We all know from experience that some practices work and others do not, and we apply them as required.

We believe that the "lab mythology" is worth studying. If the differential characteristics of the theoretical vs. actual experiment process are research-group specific, the inquiry will end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to the improvement of experimental practice.

We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring some issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic, because they require too much time) on this topic soon.

### 2023-07-02 tex/discussion.tex
First upload
This is the first upload

\section{Discussion and related work}\label{sec-related}
This research aims to find out how experimenters work in practice. After conducting the ethnographic research, we obtained models (process, workflow, conceptual). The conceptual model connects with empirical researchers' "prominent picture" of their research area. The process and workflow models represent a novel contribution; to our knowledge, such detailed descriptions of the experimental process are not available in the existing literature.

The models have been described in Section~\ref{sec-results} and are available in full detail online. In this section, we would like to focus on some interesting models features:

\begin{itemize}
	\item \textbf{The experimental process is broader} than we usually recognize as "experimentation". An experiment can be characterized as a small (or not that small) research project. It includes activities related to (see this \href{https://zenodo.org/record/7105096#.YyxsxezMLUJ}{\ul{link}}):
	\begin{itemize}
		\item management (resources, research lines, etc.),
		\item negotiation (e.g., with companies), 
		\item document/artifact support, 
		\item publication, etc.
	\end{itemize}
Although all empirical researchers are aware of these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature. To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation} \textbf{do not address these activities beyond marginal notes}. Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.
	\item The ethnographic study revealed one consequence of such diversity. It is possible, albeit unlikely, that one experimenter performs all experimental activities. These activities are not only complex: they demand time. In a project (research or not), the diversity of activities is usually associated with the existence of roles with well-defined responsibilities. \textbf{Also in experimentation. We identified three roles}: (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}. These roles are probably related to the RGUS' size; smaller (or larger) groups may have more generic/specialized roles. The concept of~\textquotedblleft role\textquotedblright~applied to experimentation is unusual but not entirely new. Mohamed et al. \cite{Mohamed-1993-roles-ESE} proposed a framework to conduct ESE experiments. This framework was based on a merge between statistical and SE process concepts, \textbf{including explicit roles}.
	\item The high-level conceptual model (see this \href{https://zenodo.org/record/7102405#.YyxvFOzMLUK}{\ul{link}}) could be endorsed by any experimenter (with some complaints depending on their specialization area, as it happened during the ethnography). However, when the concept of role appears in the scene, the high-level model is just \textbf{an abstraction of the information handled by the roles}. In other words: the high-level model is incomplete from the roles' viewpoint and thus partially valid only. The high-level model evokes the viewpoint of the \textit{senior experimenter} (see this \href{https://zenodo.org/record/7102464#.YyxvsezMLUK}{\ul{link}}), but with fewer details. In turn, the \textit{experiment manager} (see this \href{https://zenodo.org/record/7102450#.Yyxv6ezMLUL}{\ul{link}}) and the \textit{research manager} (see this \href{https://zenodo.org/record/7102431#.YyxwGezMLUL}{\ul{link}}) pay little attention to the design and execution details because the \textit{senior experimenter} takes care of that. Likewise, the \textit{experiment manager} is not affected by the experimental research in its global context, e.g., collaboration with other groups and replication. Such a concern fits the \textit{research manager}'s role. 
\item We deliberately built the models (process, workflow, and conceptual) to offer a coherent picture. However, as we indicated in Section~\ref{subsubsec-focus-groups}, we could not address the existence of different families of experiments. When experimenters working in other families get together, the models automatically diverge. Therefore, the models represent the maximum achievable consensus obtained during the role-oriented meetings. We are conscious that \textbf{further research is necessary at the family level to discover the peculiarities of each research area}. Such an inquiry could have an impact on the SE community. 
\item The replication problem is related to knowledge transfer \cite{Shull-2004-Knowledge-sharing-issues-SE}. Nevertheless, we wonder what exactly does it mean to transfer knowledge? To what extent or how should it be done? For example, Ferreira et al. \cite{Ferreira-2017-planning-experiments} emphasize methodological consistency over discussing methods and techniques.

There is, apparently, an underlying dichotomy between theory and practice. In theory, formal guides, e.g., replication packages, should describe the process step by step. However, the practice shows that such mechanisms cannot cope with the existing diversity of experimental protocols. In our opinion, the replication process will likely benefit from the specialized, fine-grained knowledge \textbf{at the family level}.
\end{itemize}

The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement. Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE}, \cite{Sjoberg-2005-survey-experiments-SE}, \cite{Shull-2004-Knowledge-sharing-issues-SE}. We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character. Instead, we aimed to observe how experimenters work, and we found out that they perform tasks other than the ones portrayed in textbooks. \textbf{This is our essential contribution}.

We are also aware that models represent \textbf{the point of view of a single experimental research group}. We must extend the inquiry to other experimental research groups, ideally to large portions of the ESE community, and check whether they reveal similar behavior patterns.

### 2023-07-02 tex/introduction.tex
First upload
This is the first upload

\section{Introduction}\label{sec-introduction}
Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives} \cite{Baker-2016-lid-reproducibility}. In Empirical Software Engineering (ESE), the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.

Several reasons prevent replication. They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE}. In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.

Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments. Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE} \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data-generation processes vary.

This paper aims to find out \textbf{how experimental researchers conduct experiments in practice}. We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities that they perform on a day-to-day basis. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different sub-areas or families of experiments have about the overall process.

The contributions of this paper are:

\begin{itemize}
  \item Identifying activities that depart from, or are not described in, mainstream experimentation procedures. 
  \item Characterizing roles, i.e., groups of experimental activities performed by the same person, in ESE research.
  \item Evidence of the existence of tacit knowledge in ESE, which, in turn, leads to discrepancies between experimentation protocols and actual practice.
  \item Conceptual and process models that represent how experimentation is conducted \textit{in practice} in an experimental research group.
\end{itemize}

Finally, we wish to open a discussion in the ESE community: 
\begin{itemize}
  \item Should those differential characteristics in experimentation in software engineering be disregarded as inherently contextual?
  \item  \textbf{Alternatively}: Are they more common than it may seem, so they should be further investigated and eventually integrated into mainstream practice?
\end{itemize}

The remainder of the article is structured as follows: Section \ref{sec-research-method} describes the research method. Section \ref{sec-reseach-execution} reports the stages of the ethnographic research and the observations made. Section \ref{sec-results} reports the concept and process models. Section \ref{sec-related} discusses the results and the connections with the existing literature. Section \ref{sec-threats} details the threats to validity and the related mitigation strategies. Finally, in section \ref{sec-conclusions}, we present the conclusions and possible future work.

### 2023-07-02 tex/research-execution.tex
First upload
This is the first upload

\section{Ethnographic study}\label{sec-reseach-execution}
We report the results of the ethnographic study in chronological order: First, we describe the analysis of the RGUS's preferred literature sources and the review of the RGUS' specific experimental material. Later, we comment on our experiences replicating an existing experiment. Finally, we describe how we formalized the information into concept and process models using interviews, focus groups, and workshops.

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support} \cite{Strandberg-2019-Ethical-Interviews-SE} \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers started conducting interviews to clarify the RGUS' knowledge and convert it directly into models.

The interviews with RGUS members provided only public, limited knowledge about experimentation. The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}. We did study the literature.

However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments. For example, the experimenters supported their story-telling in specific experimental materials that the researchers did not have. Nor did the researchers have the knowledge required to understand the experimenters' discussions. At this point, the need for a different research methodology --ethnography-- became apparent. Consequently, the researchers switched minds and continued at a slower pace. We carried out additional activities to understand the standard literature and experimental material. The interaction with the experimenters also changed. For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS families of experiments.

\subsubsection{Analysis of the relevant literature}
The experimenters use the books by Wohlin et al. \cite{Wohlin-2000-Experimentation-SE}, and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process. They also utilize specific papers that advise how to experiment in SE, e.g., Basili et al. \cite{Basili-1986-ESE}, and Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}. Basili et al.'s work \cite{Basili-1999-families-experiments} was also considered relevant due to the experimenters' interest in carrying out experimental replications.

We reviewed the literature in-depth. We noticed that the sources do not use coherent terminology. At the outset, it seemed to the researchers that the literature describes different experimental procedures. They later verified that the different terms refer to essentially the same activities, e.g.:

\begin{itemize}
	\item Wohlin et al. \cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report".
	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution and Analysis".
	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation".
	\item Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data collection, Analysis, Presentation of results, and Interpretation of results".
\end{itemize}

We looked for additional ESE books to verify whether terminological diversity was a widespread problem. We only found a new edition of Wohlin et al.'s book \cite {Wohlin-2012-experimentatio-SE}, and the specialized books by Runeson et al. \cite{Runenson-2012-case-study-SE} and Kitchenham et al. \cite{Kitchenham-2015-Evidence-Based-SE}. These books share standard terms, as expected, but they use diverse terms throughout the text, contributing to the terminological diversity.

\subsubsection{Review of experimental material}\label{subsubsec-review-experimental-material}
The RGUS' experimental materials are a promising source of information since they represent the work performed during almost two decades of experimentation and reflect the tacit knowledge \cite{Polanyi-1996-tacit-k} present in the group. However, accessing the experimental materials was challenging; materials were scattered and disorganized. We needed the experimenters' help to find specific documents/data files, understand their structure and contents, and gain further access to other related materials.

The experimental materials evolve. Experiments vary widely in documentary support, independently of their date. The experimental family explains the differences better than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.

Concerning the experimental process, all experimental materials describe the activities carried out in a general way. The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge. Terminological diversity also affects experimental materials, especially among families of experiments.

We intended to reflect the RGUS knowledge in process and conceptual models. However, we largely failed to consolidate the terminology, so the resultant model, e.g., see this 
\href{https://zenodo.org/record/7093417#.YyjHb-zMLUI}{\ul{link}}, was relatively simple. However, this abstract preliminary model was satisfactory for continuing the ethnographic research.

\subsection{Becoming active participants}\label{replication}
The concepts acquired by the researchers were uncertain. We needed some external confirmation. We decided to apply the learn-by-doing methodology \cite{mendoza-2019-learn-by-doing-methodology} by experimenting and determining whether our understanding was correct.

\subsubsection{Experiment selection}
Instead of designing an experiment from scratch, the researchers opted to replicate an experiment previously conducted by the RGUS. The replication has been published elsewhere.

Conducting a replication of the experimenters' previous work attracted their attention and guaranteed collaboration. Furthermore, the replication would give us the chance to recreate the design, execution, analysis, and reporting of an experiment that, until then, we only knew due to the examination of the written materials. Finally, the replication would help us better understand the experimenters' motivations and decisions.

\subsubsection{Replication planning}
Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE} \cite{Vegas-2006-communication-researchers} \cite{Miller-2005-replicating-SE-experiments} \cite{Gomez-2014-understanding-replication} \cite{Demagalhaes-2015-replications-SE} \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities. For example, how to approach the experimental design, which experimental objects are necessary, and in which order. We requested the cooperation of the experimenters who ran the baseline experiment to schedule the experimental activities and adapt the design to the new context where the replication would be run.

\subsubsection{Conduction}
Efra\'in R. Fonseca C. performed a literal replication \cite{Gomez-2014-understanding-replication} in an external academic setting to the RGUS.

The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined in advance all relevant aspects. The measurement was the most complex and controversial aspect. Let's say that the experimental subjects return test cases, but these test cases are manually interpreted and not run. The experimenters' cooperation was again needed to complete the measurement. Even if the measurement procedure looked controversial, it made complete sense after discussion. We learned that the experimental protocols depend on the experimental task, which in turn depends on the research area. The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty. It was possible to reproduce the duplicate files used in the baseline experiment.

\subsubsection{Analysis and reporting}
We sought help to analyze and report the replication since we did not have statistical training. In retrospect, both the analysis and the report were simple to perform since the experimental design used (\textit{cross-over}) was associated with a well-defined analysis method \cite{Vegas-2016-crossover-designs-experiments-SE}. On the other hand, there were reporting guidelines available \cite{Carver-2010-guidelines-replication-SE}. The experimenters confirmed that, generally, the analysis and reporting of experiments do not imply substantial SE challenges.

The differences observed between replication and the baseline experiment confirmed our belief that, at least in the case of the RGUS, there are no clear guidelines that allow literal, external, and independent replication \cite{Gomez-2014-understanding-replication} of the original experiment, irrespective of the existence of incidents in the replication's execution. As a corollary, the original experimenters' presence was necessary to make the baseline experiment results and the replication comparable.

\subsubsection{Assessment}
The replication was, in retrospect, the key to clarifying the RGUS' experimental process. The replication exercise increased the researchers' knowledge and enabled the verification of previous findings, e.g., the RGUS terminology. We made some discoveries, e.g., the specificity of experimental protocols and, notably, the confirmation of extensive tacit knowledge.

\subsection{Modeling the RGUS' knowledge}\label{subsec-conocimiento-grupo}
After the experiment, the researchers were not sure which way to go. The investigation was successful from an operational viewpoint, except for the analysis, which required specific statistical knowledge. The researchers satisfactorily completed all other activities. The models improved somehow during the replication, e.g., see this \href{https://zenodo.org/record/7101676#.YytEquzMLUI}{\ul{link}}, and to the best of our knowledge, we did not overlook anything relevant.

However, we were not completely satisfied with the ethnographic study. We were able to replicate \textbf{one} experiment in \textbf{one} concrete SE area, but nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}. The presence of tacit knowledge represented a severe threat to validity.

\subsubsection{Semi-structured interviews}
We decided to organize new series of semi-structured interviews with the experimenters. The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on the replication of \textit{specific families of experiments}. Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.

We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences. We gave up note-taking and captured the interviews on video, including the experimenters' body language. We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative. We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols}. We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
	\caption{Example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

The process was smooth since each experimenter offered a coherent narrative. We conducted a total of seven interviews with six RGUS members. Each time, we refined the acquired knowledge. The reader can check the models' evolution through this \href{https://zenodo.org/record/7102213#.YytmMezMLUI}{\ul{link}}. When the progress was marginal, the interviews concluded.

\subsubsection{Preliminary models}\label{sec:preliminary-models}
We created process and concept models. We also created a workflow to represent the relations between activities that appeared during the investigation. Please notice that we introduce the outstanding characteristics of the models only because the ethnographic research did not end at this point yet. The models and the workflow have to be validated by the experimenters. We will provide full details in Section~\ref{sec-results}.

\begin{itemize}
	\item \textit{Preliminary} experimental process model: We realized that the way the RGUS conducts experiments resembles the processes we are used to in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}. In the end, they all are processes. Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization. Each process has several related activities. For example, the \textit{Basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).

\item \textit{Preliminary} workflow: It shows how the connections between experimental processes (see this \href{https://zenodo.org/record/7102360#.Yyt1a-zMLUI}{\ul{link}}). According to the experimenters, the experimental process has three main stages or paths: experimentation, replication, and synthesis. The activities performed in each step can be classified according to the experimental process model described above.

An experimenter does not usually carry out all activities, although they know how to perform them (maybe not flawlessly). In turn, different responsibilities are typically assigned to the same experimenter. We identified the following roles (the names suggest their responsibilities): (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}. These roles are responsible for: (1) planning the experimental research, (2) managing the logistics, and (3) conducting the experiment in practice, respectively. The study showed overlapping responsibilities among roles, e.g., the domain and topic selection for the experiment.

	\item \textit{Preliminary} conceptual model: Preliminary results at the conceptual model level show activities similar to what the books say, indicating that it has evolved the least (see this \href{https://zenodo.org/record/7102387#.Yyt7W-zMLUI}{\ul{link}}).
\end{itemize}

\subsubsection{Double-checking using focus groups}\label{subsubsec-focus-groups}
Interviews have limitations for eliciting tacit knowledge. When the preliminary models were ready, we made a presentation to all RGUS experimenters for double-checking.

As we said above, we organized the semi-structured interviews around families of experiments. It was a fortunate mistake. We found out that the families of experiments are pretty homogeneous. The experimenters involved in a family quickly agree and provide a coherent picture of the experimental process. However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families. For instance, the experimenters never agreed on the activities related to the experimental design.

We aimed to solve the disagreements using a consensus-reaching technique, e.g., DELPHI \cite{Dalkey-1967-Delphi}. However, social pressure was not a problem. We decided to organize focus groups. The most senior experimenter led the meetings as a moderator, but only to keep the discussion on track. The researchers only observed. Each session lasted at least two hours. Before the focus groups, the experimenters reviewed the concept models. We started with the concept models because they were more straightforward and smaller. The process models/workflow were later updated based on the changes to the concept model.

The focus groups prompted a substantial growth in the models' details after heated discussions that re-confirmed the existence of terminological diversity \textit{even among experimenters who carry out similar activities within the same family of experiments}. The most likely reason seems to be the training that each experimenter received. Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions. The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches} \cite{montgomery-2019-Design-Analysis-Experiments} did not drive self-education. Few of these materials existed when the experimenters started their careers. The experimenters chose the training materials according to their preferences and the experimental families in which they participated. Terminological diversity was the logical consequence.

The experimenters did not reach a consensus during the focus group exercise. The discussions with experimenters of different families exhibited a characteristic circular pattern: An experimenter triggered a change that, later, was undone by another. Substantial agreement was only possible \textit{within each family of experiments}. The progress we achieved during the replication phase of the ethnographic research (see Section~\ref {replication}) was possible due to our focus on concrete families of experiments, a fact that we completely ignored at that time.

The concept model that we obtained at this stage appears in this \href{https://zenodo.org/record/7102405#.YyuAfOzMLUI}{\ul{link}}. Notice that the agreement is apparent.

\subsubsection{Role-focused workshops}\label{subsubsec-focus-groups-role}
We did not aim to explore SE family-specific experimental processes. Albeit enjoyable, we perceived it as future research. In turn, we tried to take advantage of the limited PhD time available to improve the models by switching the perspective from experimental families to \textit{experimental roles}. There was no reason to think that roles could contribute to the agreement. However, we used the concept model to drive the focus groups, not the process models. Roles are connected with activities, not concepts. Activities looked more uniform than concepts. It was worth trying.

The strategy was simple. Experimenters who played the same role gathered informally and discussed the tasks associated with that role \textit{only}. Meetings were peaceful (the focus group meetings were less gentle), and the conversations were constructive. Putting aside family-specific issues, we could double-check the existence of roles (the experimenters agreed that roles make sense, even if they have never perceived their responsibilities in these terms).

We could also break down the conceptual model into role-specific conceptual models. They represent groups of concepts used by the experimenters when they play a specific role. Most experimental activities are role-specific, e.g., the senior experimenter role uses the run-kit concept exclusively. The role-specific conceptual models are available through these links: \href{https://zenodo.org/record/7102431#.YyuFvuzMLUI}{\ul{1}}, \href{https://zenodo.org/record/7102450#.YyuG2OzMLUI}{\ul{2}}, \href{https://zenodo.org/record/7102464#.YyuIkOzMLUI}{\ul{3}}.

### 2023-07-02 tex/research-method.tex
First upload
This is the first upload

\section{Research Method}\label{sec-research-method}
Ethnography is a \textit{participant observation} method. It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}. Ethnographic researchers adopt "the ESE researchers' point of view" to better understand the reason for their actions \cite[pp. 55-56]{Outhwaite-2007-sage}. Thus, the biases and misperceptions that ESE researchers might have about themselves can be detected. Ethnography also captures the rich context that surrounds ESE researchers.

Ethnography identifies details that are ignored with other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted for the study of a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

We followed Runenson et al.'s case studies guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:

\begin{itemize}
	\item \textbf{Context selection}, i.e., choose the experimental research group that facilitates the research in the best possible condition.
    \item \textbf{Data collection procedures}.
    \item \textbf{Data analysis procedures}.
\end{itemize}

To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters". We will talk about ourselves the authors as the "the researchers". We the authors/researchers play two different roles in the investigation:

\begin{itemize}
\item \textbf{Efra\'in R. Fonseca C.} and Edison G. Espinosa conducted the ethnographic study as part of their Ph.D. Edison G. Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D. This is the reason why he is not authoring this paper. The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R. Fonseca C.
\item \textbf{Oscar Dieste} and \textbf{Natalia Juristo} supervised the investigation. Their views and opinions are not reflected in this manuscript.
\end{itemize}

\subsection{Context selection}
We have considered an experimental research group as "representative", i.e., a reliable source of information, when it fulfills the following selection criteria:

\begin{itemize}
	\item The experimental research group has extensive experience in ESE research.
	\item The ESE community recognizes the experimenters as specialists in their areas.
	\item The group has a large number of publications in journals of repute and internal documentation, i.e., the group has a verifiable track record.
\item One of the main drivers of this research is to explore the connection between replication and lab practices. We positively value that the experimental research group conducts replications or families of experiments.
\end{itemize}

Based on these criteria, we selected an experimental research group with more than 20 years of experience in experimentation in SE. The experimenters have a long track record of publication in journals and conferences of repute. The group has focused on replications and families of experiments since they consider that isolated experiments do not contribute significantly to scientific knowledge. The manuscript conceals the group's identity but is disclosed in the raw data repository (see Section \ref{sub-sec-research-documentaition}).

\subsection{Data collection procedures}\label{subsec-data-collection}

The information sources within the experimental research group under study (RGUS) are:

\begin{itemize}

	\item Experimenters' knowledge.

	\item General and specific literature, e.g., the book \textquotedblleft Experimentation in Software Engineering\textquotedblright~\cite{Wohlin-2012-experimentatio-SE}.

	\item General and specific experimental materials, e.g., experimental designs.

	\item Periodic and/or \textit{ad-hoc} activities carried out, e.g., experimental planning.

\end{itemize}

Ethnography information gathering is characterized by being cyclical, iterative, and incremental. The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information. Such taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}. Table \ref{tbl-tecnica-fuente} describes the information-gathering techniques and instruments and the corresponding information sources to which they can be applied.

\begin{table*}
	\small
	\centering
	\caption{Techniques and tools for information acquisition, by source}
	\label{tbl-tecnica-fuente}
	\begin{tabular}{|p{2cm}|p{3.5cm}|p{5cm}|}
	\hline
	\textbf{Source} & \textbf{Technique} & \textbf{Instruments}\\
	\hline
	\multirow{7}{50 pt}{Experimenters knowledge} & Interviews & Video camera, paper, pencil\\
	\cline{2-2}\cline{3-3}
	& Focus groups & Video camera, blackboard, cardboard, board markers, chalkboard eraser, etc.\\
	\cline{2-2}\cline{3-3}
	& Process modeling & Video camera, paper, pencil\\
	\cline{2-2}\cline{3-3}
	& Instrumental systems & Email, videoconferences, internet\\
	\cline{2-2}\cline{3-3}
	& Experience-based learning & Literature review, review of experimental material, experimentation in practice\\
	\hline
	\multirow{3}{15 pt}{Common literature} & Documentation analysis & Comprehensive reading of literature, reading summaries, software support tools\\
	\hline
	\multirow{2}{15 pt}{Experimental material} & Documentation analysis & Systematic analysis of experimental material, trials with data\\
	\hline
	\multirow{2}{15 pt}{Group activities} & Participatory observation & Literature review, researchers knowledge\\
	\hline
	\end{tabular}
\end{table*}

The order in which researchers contact information sources is essentially random. The frequency of contact depends on the researchers' needs to answer research questions during the ethnographic research and the availability of each source.

\subsection{Data analysis procedure}

Ethnographic practice involves storing the raw data and then interpreting the findings \cite{Cooper-2007-Sharing-Data-Ethnographic-Research}. In this research, the data analysis procedure consists of four activities:

\begin{itemize}
\item \textbf{Preparation of materials}: The information sources are classified and categorized so experimenters and researchers can manage them.
\item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
\item \textbf{Verification of information}: This is probably the most complex activity. The knowledge obtained from the sources complements each other. We will use triangulation to explain the findings and identify new areas of inquiry. All findings are discussed with the experimenters to avoid misunderstandings and verify that the resulting activities are those that they carry out in practice.
\item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters. The expected results will be models (conceptual and process) and the practices derived from the interviews with the experimenters.
\end{itemize}

\subsection{Research documentation}\label{sub-sec-research-documentaition}
The raw data obtained in this study is organized as shown in Figure \ref{fig-info-study}. The tree represents a repository of around 20GB of information. The repository contains pictures, recordings, models, reports, and transcriptions, most in Spanish (the operational language used during the research). Spanish is widespread enough to utilize the repository with little effort. The final products, e.g., models, have been translated into English to improve the reader's understanding. The repository is not public because it contains personal data. However, the corresponding author will share the repository with interested researchers upon request.

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Intermediate-Products}
	\caption{Repository structure}
	\label{fig-info-study}
\end{figure}

### 2023-07-02 tex/results.tex
First upload
This is the first upload

\section{Results}\label{sec-results}
The ethnographic research yielded three main results: (1) an \textit{experimental process workflow}, (2) an \textit{experimental research concept model}, and to a lesser extent, (3) an \textit{experimental process model}. They are described below. We have not used standard notations, e.g., UML or BPMN because experimental researchers (and probably readers) have different ages, backgrounds, and careers. We have valued communication above model formality.

\subsection{Experimental process workflow}

The so-called \textit{experimental process workflow} (EPW) represents the tasks carried out during SE experimentation (according to the RGUS' perspective) and their logical sequence. The workflow is available in this \href{https://zenodo.org/record/7102486#.YyuKruzMLUI}{\ul{link}} and, with lower resolution, it is shown in Fig.~\ref{fig-EPM-construction}.

The EPW model shows three well-defined paths corresponding to the key SE experimental research processes: (1) Experimentation (purple), (2) replication (green), and (3) synthesis (yellow). In addition, based on the existence (or not) of new moderator variables and new knowledge generated in the last round, there are activities of analysis associated with planning new cycles of experimentation, replication, or synthesis and reporting related to the publication of results, respectively (highlighted in blue). Such us processes appear associated with the experimentation and synthesis paths.

\subsubsection{Experimentation path}
Experimentation is the heart of experimental research since replication and synthesis stem from it. However, in some circumstances, experimentation may require the previous execution of synthesis (e.g., when choosing the topic of the experiment) or replication (e.g., in adaptation to the context) activities. According to the experimenters, the experimentation activities are:
\begin{itemize}
	\item Selecting the experiment domain.
	\item Selecting the experiment topic.
	\item Selecting the experimental hypothesis.
	\item Planning the experiment.
	\item Creating the experimental kit.
	\item Executing and post-execution of the experiment.
\end{itemize}

When the experiment results do not meet the experimenters' expectations, they plan a new cycle of experimental research redesigning and conducting the experiment again. This redesign activity usually happens when investigating fundamental topics.

\subsubsection{Replication path}
Replication is the practice on which scientific knowledge is based, hand in hand with the synthesis of results \cite{Roizer-2014-reproducibility}. The RGUS agrees that replication is essential for experimental research both explicitly (RUGS' experimenters claim that) and, more importantly, implicitly (the experimental workflow contains minute details about replication activities). The replication path consists of the following activities: 
\begin{itemize}
	\item Definition of replication.
	\item Study and update the experimental kit.
	\item Adaptation to the context, and
	\item Creation of operational elements.
\end{itemize}

\subsubsection{Synthesis path} 
According to the RGUS experimenters, the synthesis of results can start from systematic literature reviews, systematic mapping studies, or experiments sharing a common goal (a family of experiments). In each case, the experimenters apply particular synthesis techniques (e.g., aggregated or individual patient meta-analysis), followed by interpreting the results. During the interpretation, the researchers can identify moderator variables or generalize pieces of knowledge. Such findings may lead to a reformulation of the synthesis' goal or the conduction of a new cycle of experimental research. In addition, the experimenters generally publish results after a synthesis, hand in hand with planning the next round.

The information obtained from the RGUS looks quite general and in line with the community's understanding of synthesis. That is correct. Our findings are somewhat predictable because the RGUS did not have any experimenter with the required expert knowledge available during this research. On the other hand, the limited insight into synthesis probably represents an indirect confirmation of the ethnographic research's success. Our understanding goes in proportion to the expert knowledge available, and there is a lot of specialized knowledge on the RGUS about experimentation and replication, which was our primary goal.

\subsection{Experimental research concept model}
This research's conceptual model is not monolithic; on the contrary, it is composed of three sub-models (more accurately, viewpoints) aligned with the prominent experimental roles (research manager - RM, experiment manager - EM, and senior experimenter - SrEx). The activities described in the EPW lead to such views. The experimenters took into account the best-known generic phases of the experimentation process to develop the new versions of the conceptual model (starting from version eight, generated in the first phase of interviews; see \href{https://zenodo.org/record/7102387#.Yyt7W-zMLUI}{\ul{link}}).

\subsubsection{Research manager (RM) viewpoint}
The RM viewpoint is available in this \href{https://zenodo.org/record/7102431#.Yyxi1ezMLUJ}{\ul{link}}. The RM has the following perspective:
\begin{itemize}
	\item In the experiment definition phase, the RM manages the "problem definition" and the "hypothesis definition." 
	\item During the experiment design phase, the RM takes care of the "human resource management" that could "run," "replicate," or be interested in "replicating" the experiment. In addition, the RM receives information regarding the "Expected results," "Execution context," and the "new variables of interest."
	\item While the experiment runs, the RM monitors events. However, the data acquisition phase is transparent to her.
	\item During the data analysis phase, the RM learns the "analysis results" and contributes to the definition of the "findings," "publications," and "pieces of knowledge."
\end{itemize}

\subsubsection{Experiment manager (EM) viewpoint}
The EM viewpoint is available in this \href{https://zenodo.org/record/7102450#.YyxjmezMLUJ}{\ul{link}}. The perspective is as follows:

\begin{itemize}
	\item In the experiment definition phase, the EM actively interacts with all experiment entities, hand in hand with the Senior Experimenter (SrEx). However, in the "hypothesis definition," the EM is only an observer.
	\item In the experiment design phase, the EM focuses on improving the "experiment guides" regarding "instruments," "objects," and the "experimental kit." Additionally, the EM can analyze the context in which the experiment occurs.
	\item During the experiment execution and data acquisition phases, the EM does not lose sight of the "events" that arise and could affect the results. The EM also analyzes the "correctness" of the "collected items" and the "raw data."
	\item In the data analysis phase, the EM cooperates with the "re-analysis" in search of new "variables of interest" to the RGUS. Additionally, the EM can observe the analysis results to suggest findings that could be useful in the publications.
\end{itemize}

\subsubsection{Senior experimenter (SrEx) viewpoint}
The SrEx viewpoint is available in this \href{https://zenodo.org/record/7102464#.Yyxl4ezMLUJ}{\ul{link}}. The perspective is as follows:

\begin{itemize}
	\item In the experiment definition phase, the SrEx observes the "experiment definition," hypotheses' "levels," and "metrics" in the first phase of the experimental investigation. The SrEx participates in the definition of the "hypothesis," "factors," and the "response variable."
	\item In the experimental design phase, the SrEx interacts with a large number of entities; hence, it was necessary to divide such a phase into three sub-phases for this role: \textit{experimental operation design}, \textit{experiment design}, and \textit{artifacts generation}.
    \begin{itemize}
	    \item During the \textit{experimental operation design}, the SrEx defines the "experimental objects" for each "experimental task," considering the minute details of each experimental object's characteristics.
	    \item During the \textit{experiment design} proper, the SrEx establishes the execution context's "parameters," the "design type" with its "groups" and assigned "experimental subjects," the "levels" defined and their combination per "session" and "period," "experimental instruments," and the "measurement procedure." 
        \item Finally, in the \textit{artifact generation}, the SrEx fundamentally develops the "run-kit," which includes the construction of the instances of the "guides," "experimental instruments," and "experimental objects." Such "instances" are the tools that the "subjects" need to participate in the "experiment."
	\end{itemize}
	\item While the experiment runs, the SrEx records all the events produced to analyze their impact on the experiment results.
	\item During data acquisition, the SrEx records the resulting collected items, creates the raw data and obtains measurements.
	\item Finally, in the data analysis phase, the SrEx performs the analysis, obtains the findings, and prepares the publications with the RM and EM's help.
\end{itemize}

\subsection{Experimental process model}
The workflow and the conceptual models represent the experimental research process carried out in the RGUS. Other experimental research groups may work differently. We have created an initial (generic, i.e., generally applicable) experimental process model (EPM) with the help of the most experienced experimenter at RGUS. The model is inspired by the software life cycle \cite{ISO-IEC-IEEE-12207}. We have adapted the concepts used in the ISO/IEC 12207 to the SE experimentation realm. Most of the items have been taken from the EPW (see Figure \ref{fig-EPM-construction}); a lesser portion derives from the conceptual models. The complete process model can be obtained in this \href{https://zenodo.org/record/7105096#.YyxoCOzMLUI}{\ul{link}}.

The EPM contains six groups of processes that define the SE experimental research life cycle: (1) Basic process, (2) Support process, (3) Knowledge item generation process, (4) Publication process, (5) Synthesis process, and (6) Organizational process.

\subsubsection{Basic process}
The basic process involve fundamental activities that experimenters perform as RM, SrEx, and EM roles before, during, and after running an experiment. Before the experiment, the activities carried out focus on its design, e.g., adapting the experiment to the research context. During the experiment, experimenters pay attention to, e.g., the events that occur before, during, and after the experiment's execution session. The activities performed after the execution focus on the data, e.g., generating raw data and data analysis.

\subsubsection{Support process}
The support process includes activities that the EM performs to manage the materials and support those interested in performing replications, as:

\begin{itemize}
	\item The management of materials before the experiment's execution (e.g., experiment design document), 
	\item the materials used during the experiment (e.g., templates for data collection), 
	\item the materials obtained at the end of the experiment (e.g., filled templates and subjects' code.), and
	\item the Knowledge transference as support style during replication activities.
\end{itemize}

\subsubsection{Knowledge items generation process}
The knowledge items generation process consolidate activities carried out by the RM to manage the knowledge acquired after each experiment, e.g., the knowledge generated in each experimentation phase, the knowledge transferred to professionals, or the knowledge generated during the identification of moderator variables.

\subsubsection{Publications process}
This group of activities focuses on reporting the experiment findings (led by the SrEx), aggregation of results (by the EM), and experimentation goals (by the RM).

\subsubsection{Synthesis process}
The EM carries out the synthesis processes. There are different types of synthesis with different evidential levels, e.g., meta-analysis, re-analysis, and informal aggregations.

\subsubsection{Organizational process}
The organizational process include RM's management activities over the SE experimental research cycle, e.g., human resources and publications management. Likewise, the EM also performs management activities on the experiment materials. Finally, there are activities related to synthesis management that have not been assigned to any role yet.

\begin{figure*}[htbp]
\begin{center}
\includegraphics[trim=0 0 0 48,clip,width=0.95\textwidth]{images/Experimentation-Process-Model-Construction}
\caption{Relationships between the \textit{experimental process workflow} (top) and the \textit{experimental process model} (down).}
\label{fig-EPM-construction}
\end{center}
\end{figure*}

Figure \ref{fig-EPM-construction} shows a not very good coincidence between the activities of the EPW and those of the EPM since, as indicated before, each model has its purpose; still, we found several relationships.

Though this model is plausible, it is based on the RGUS' perspective alone. Further research (in this regard, see Section~\ref{sec-conclusions}) is necessary to ensure that the model represents the experimental SE community's experimental process.

### 2023-07-02 tex/threats-to-validity.tex
First upload
This is the first upload

\section{Threats to Validity}\label{sec-threats}
Validity is the degree to which the research is conducted transparently and without biases. We have followed Runeson et al.'s recommendations \cite[p. 71–73]{Runenson-2012-case-study-SE} to design the ethnographic research and describe the threats to validity. Runeson et al., following Yin \cite{Yin-2009-case-study}, classify threats in four groups: (1) construct validity, (2) internal validity, (3) external validity, and (4) reliability.

\subsection{Construct validity}
Construct validity deals with the degree of agreement between the study constructs, i.e., what the ethnographic researchers have in mind, and the observations made in the field. For example, if the interview questions are not interpreted the same way by the researcher and the experimenters, there is a threat to construct validity.

The ethnographic method is particularly suitable to oppose this type of threat because knowledge acquisition progress at a slow pace. On top of that, we used several techniques during the research to create constructs, e.g., reading, observation, interviews, among others. The different approaches increase the chances of identifying or solving misunderstandings between the researchers and the RGUS. We also prevented construct validity threats using reliable information sources, e.g., standard literature and the RGUS' experimental material.

Finally, the constructs were validated iteratively by experimenters and researchers during the research project lifespan.

\subsection{Internal validity}
Internal validity addresses the credibility of the causal relationships found during the research. However, this threat does not apply to this ethnographic study, as we do not aim to identify causal relationships but, at most, correlations between phenomena, e.g., when we claim that the number and diversity of experimental tasks explain the existence of roles. Furthermore, most of the findings we provide are descriptive, e.g., the disagreements among experimenters of different experimental families.

\subsection{External Validity}
External validity represents the degree of certainty of the findings obtained in an investigation being generalizable, i.e., applied to different contexts of their origin. This threat is prominent in this study since it was carried out in a single instance (an experimental research group) of the population under study (SE research groups that apply empirical methods). As a countermeasure, we chose a highly representative experimental research group.

\subsection{Reliability}
The reliability of a study is the ease with which the research activities and the results obtained can be reproduced by other studies that apply the same methodology in the same RGUS (more likely, in similar experimental research groups).

As mitigation measures, we have followed the research methodology outlined in Section~\ref{sec-research-method} closely to ensure the reliability of this research. Moreover, we have disclosed all intermediate and final results in a repository, available under request, to guarantee transparency in the investigation.

### 2023-07-04 tex/abstract.tex
Abstract resumido

@@
 \begin{abstract}
-\textbf{Context:} Experimental replication is complex in Software Engineering.
-Several strategies, such as replication packages and families of experiments, have been applied with limited success.
-We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, etc., or intrinsic reasons, i.e., the researchers' needs are not fulfilled by the proposed replication procedures.
-\textbf{Objective:} To find out how ESE researchers conduct experiments in practice.
-\textbf{Method:} We carried out an ethnographical study with an experimental research group for two years instead of working halfway with several research groups over short periods.
-\textbf{Results:} We have created conceptual and process models representing how experimentation, replication and synthesis are conducted in the target research group.
+\textbf{Context:} Replication is complex in Experimental Software Engineering (ESE).
+Replication packages and families of experiments have been applied with limited success.
+We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, etc., or intrinsic reasons, i.e., the researchers' needs are not fulfilled by the existing replication procedures.
+\textbf{Objective:} Find out how ESE researchers conduct experiments in practice.
+\textbf{Method:} Ethnographical study with an experimental research group.
+\textbf{Results:} We have created conceptual and process models representing how experimentation, replication, and synthesis are conducted in the target research group.
 These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels.
-The experimental process carried out in the group differs from textbooks in terms of (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different sub-areas or families of experiments have about the experimental process.
-\textbf{Conclusions:} The differences between the actual lab processes vs the idealized versions portrayed in textbooks probably harm knowledge transfer among labs and make replication harder.
-However, our observations are limited to one research group.
-We plan to conduct further research with a representative sample of the research groups, e.g., using a survey, to explore whether our findings have general applicability.[CHECK NO MORE THAN 200 WORDS]
+The experimental process carried out in the group differs from folk knowledge and textbooks in (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts, and (4) the viewpoints that different sub-areas or families of experiments have about the experimental process.
+\textbf{Conclusions:} The differences between the actual lab processes vs.
+the idealized versions probably harm knowledge transfer and make replication harder.
+We will conduct further research with more research groups, e.g., using a survey, to explore whether our findings have general applicability.
 \end{abstract}

### 2023-07-23 tex/abstract.tex
some more changes
A couple of more changes.

@@
 Replication packages and families of experiments have been applied with limited success.
-We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, etc., or intrinsic reasons, i.e., the researchers' needs are not fulfilled by the existing replication procedures.
+We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, or intrinsic reasons, i.e., the existing replication procedures do not fulfill the researchers' needs.
 \textbf{Objective:} Find out how ESE researchers conduct experiments in practice.
 \textbf{Method:} Ethnographical study with an experimental research group.
-\textbf{Results:} We have created conceptual and process models representing how experimentation, replication, and synthesis are conducted in the target research group.
+\textbf{Results:} We have created conceptual and process models representing experimentation, replication, and synthesis in the target research group.
 These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels.
@@
 the idealized versions probably harm knowledge transfer and make replication harder.
-We will conduct further research with more research groups, e.g., using a survey, to explore whether our findings have general applicability.
+We will conduct further research with more research groups, e.g., using a survey to explore whether our findings have general applicability.
 \end{abstract}

### 2023-07-23 tex/introduction.tex
some more changes
A couple of more changes.

@@
 We observed the researchers in their lab for two years.
-We created conceptual and process models to represent the concepts they use and the activities that they perform on a day-to-day basis.
+We created conceptual and process models to represent the concepts they use and the activities they perform daily.
 These models fit the community's procedures and terminology at a high level.
@@
 \begin{itemize}
-  \item Identifying activities that depart from, or are not described in, mainstream experimentation procedures. 
+  \item Identifying activities that depart from or are not described in mainstream experimentation procedures. 
   \item Characterizing roles, i.e., groups of experimental activities performed by the same person, in ESE research.
@@
   \item Should those differential characteristics in experimentation in software engineering be disregarded as inherently contextual?
-  \item  \textbf{Alternatively}: Are they more common than it may seem, so they should be further investigated and eventually integrated into mainstream practice?
+  \item  \textbf{Alternatively}: Are they more common than they may seem, so they should be further investigated and eventually integrated into mainstream practice?
 \end{itemize}

### 2023-07-24 tex/research-method.tex
Last changes
Last changes  made before the submission.

@@
 
-\begin{table*}
+\begin{table}
 	\small
@@
 	\label{tbl-tecnica-fuente}
-	\begin{tabular}{|p{2cm}|p{3.5cm}|p{5cm}|}
+	\begin{tabular}{|p{2cm}|p{2.5cm}|p{3cm}|}
 	\hline
@@
 	\hline
-	\multirow{7}{50 pt}{Experimenters knowledge} & Interviews & Video camera, paper, pencil\\
-	\cline{2-2}\cline{3-3}
+	\multirow{14}{50 pt}{Experimenters knowledge} & Interviews & Video camera, paper, pencil\\
+	\cmidrule{2-2}\cmidrule{3-3}
 	& Focus groups & Video camera, blackboard, cardboard, board markers, chalkboard eraser, etc.\\
-	\cline{2-2}\cline{3-3}
+	\cmidrule{2-2}\cmidrule{3-3}
 	& Process modeling & Video camera, paper, pencil\\
-	\cline{2-2}\cline{3-3}
+	\cmidrule{2-2}\cmidrule{3-3}
 	& Instrumental systems & Email, videoconferences, internet\\
-	\cline{2-2}\cline{3-3}
+	\cmidrule{2-2}\cmidrule{3-3}
 	& Experience-based learning & Literature review, review of experimental material, experimentation in practice\\
 	\hline
-	\multirow{3}{15 pt}{Common literature} & Documentation analysis & Comprehensive reading of literature, reading summaries, software support tools\\
+	\multirow{4}{15 pt}{Common literature} & Documentation analysis & Comprehensive reading of literature, reading summaries, software support tools\\
 	\hline
-	\multirow{2}{15 pt}{Experimental material} & Documentation analysis & Systematic analysis of experimental material, trials with data\\
+	\multirow{3}{15 pt}{Experimental material} & Documentation analysis & Systematic analysis of experimental material, trials with data\\
 	\hline
@@
 	\end{tabular}
-\end{table*}
+\end{table}
 

### 2023-07-25 Article.tex
Corregido el problema de la bibliografia y de las listas itemize
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

@@
 \usepackage{framed}
+\usepackage{graphicx}
 \usepackage{multirow}
+\usepackage{booktabs}
 \usepackage{color}
@@
 \usepackage{enumitem}
-\usepackage{hyperref}
-\usepackage{url}
+	\setlist[itemize]{leftmargin=5.5mm}
+
+%\usepackage{hyperref}
+%\usepackage{url}
+\usepackage[hyphens]{url}
+\usepackage[hidelinks]{hyperref}
+\hypersetup{breaklinks=true}
+\urlstyle{same}
+%\usepackage{cite}
 \usepackage{soul}
@@
 \input{tex/conclusion}
-\input{tex/data-availability-statements}
+%\input{tex/data-availability-statements}
 %%%%%%% End Content %%%%%%%%%
@@
 \section*{Acknowledgment}
-
-Insert the Acknowledgment text here.
+This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
 
@@
 
-\bibliographystyle{imaiai}
+\Urlmuskip=0mu plus 1mu\relax
+\bibliographystyle{iet}
 \bibliography{bibliography}

### 2023-07-25 tex/abstract.tex
Last changes
These are the last changes previous the submission

@@
 We wonder whether the problems are due to formal issues, e.g., under-specification, miscommunication, or intrinsic reasons, i.e., the existing replication procedures do not fulfill the researchers' needs.
-\textbf{Objective:} Find out how ESE researchers conduct experiments in practice.
+\textbf{Objective:} Determine how ESE researchers conduct experiments in practice.
 \textbf{Method:} Ethnographical study with an experimental research group.
@@
 These models fit the mainstream procedures and terminology at a high level but depart at lower, specialized levels.
-The experimental process carried out in the group differs from folk knowledge and textbooks in (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts, and (4) the viewpoints that different sub-areas or families of experiments have about the experimental process.
+The experimental process carried out in the group differs from folk knowledge and textbooks in (1) the number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts, and (4) the viewpoints that different subareas or families of experiments have about the experimental process.
 \textbf{Conclusions:} The differences between the actual lab processes vs.

### 2023-07-25 tex/conclusion.tex
Last changes
These are the last changes previous the submission

@@
 \section{Conclusions}\label{sec-conclusions}
-Experimental researchers know that experimentation, as described in textbooks and articles, is only an approximation to the daily work in the labs and the field.
+Experimental researchers know that experimentation, as described in textbooks and articles, only approximates the daily work in the lab and the field.
 This study shows the deviations found in a concrete experimental research group.
-Such divergences do not seem arbitrary but necessary from a systemic viewpoint, e.g., each people have distinct abilities and perform different tasks in a research project.
+Such divergences do not seem arbitrary but are necessary from a systemic viewpoint, e.g., people have distinct abilities and perform different tasks in a research project.
 
 We do not claim that our findings are commonplace in the empirical community.
-They may be particular to the experimental research group in which we performed the ethnographic study.
-However, some clues evidence that peculiar behaviors are usual in the labs and the field.
+They may be particular to the experimental research group where we performed the ethnographic study.
+However, some clues evidence that peculiar behaviors are usual in the lab and the field.
 For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}.
 At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes.
-These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k} \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes.
+These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes.
 Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}.
-Our study has shown that such tacit knowledge is present in SE and manifests at a relatively superficial level, e.g., when researchers from different experimental families discuss their mental models.
+Our study has shown that such tacit knowledge is present in SE and manifests relatively superficially, e.g., when researchers from different experimental families discuss their mental models.
 
@@
 If the differential characteristics of the theoretical vs.
-actual experiment process are research-group specific, the inquiry will end.
+actual experiment process are specific to the research group, the inquiry will end.
 However, it is also possible that widespread patterns of behavior show up.
-Bringing such knowledge to the foreground could contribute to the improvement of experimental practice.
+Bringing such knowledge to the foreground could contribute to improving experimental practice.
 
@@
 Therefore, it is necessary to carry out further studies to generalize the results of the present work.
-We have completed a survey exploring some issues, e.g., terminological diversity, that surfaced during the ethnography.
+We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography.
 We have also completed a comparative study of the experimental process between the SE and an established engineering discipline.
 Both works will likely be published shortly.
-In addition, we plan to conduct more studies (probably not ethnographic, because they require too much time) on this topic soon.
+In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

### 2023-07-25 tex/discussion.tex
Last changes
These are the last changes previous the submission

@@
 This research aims to find out how experimenters work in practice.
-After conducting the ethnographic research, we obtained models (process, workflow, conceptual).
+After conducting the ethnographic research, we obtained models (process, workflow, and conceptual).
 The conceptual model connects with empirical researchers' "prominent picture" of their research area.
-The process and workflow models represent a novel contribution; to our knowledge, such detailed descriptions of the experimental process are not available in the existing literature.
+The process and workflow models represent a novel contribution; to our knowledge, such detailed descriptions of the experimental process are unavailable in the existing literature.
 
-The models have been described in Section~\ref{sec-results} and are available in full detail online. In this section, we would like to focus on some interesting models features:
+The models described in Section~\ref{sec-results} are available in full detail online. In this section, we would like to focus on some interesting model features:
 
@@
 		\begin{itemize}
-		\item management (resources, research lines, etc.),
+			\item management (resources, research lines, among others.),
 			\item negotiation (e.g., with companies), 
 			\item document/artifact support, 
-		\item publication, etc.
+			\item publication, among others.
 		\end{itemize}
-Although all empirical researchers are aware of these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature.
-To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation} \textbf{do not address these activities beyond marginal notes}.
+Although all empirical researchers know these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature.
+To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}.
 Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.
@@
 These activities are not only complex: they demand time.
-In a project (research or not), the diversity of activities is usually associated with the existence of roles with well-defined responsibilities.
+In a project (research or not), the diversity of activities is usually associated with roles with well-defined responsibilities.
 \textbf{Also in experimentation.
-We identified three roles}: (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}.
+We identified three roles}: (1) \textit{Research Manager}, (2) \textit{Experiment Manager}, and (3) \textit{Senior Experimenter}.
 These roles are probably related to the RGUS' size; smaller (or larger) groups may have more generic/specialized roles.
@@
 	\item The high-level conceptual model (see this \href{https://zenodo.org/record/7102405#.YyxvFOzMLUK}{\ul{link}}) could be endorsed by any experimenter (with some complaints depending on their specialization area, as it happened during the ethnography).
-However, when the concept of role appears in the scene, the high-level model is just \textbf{an abstraction of the information handled by the roles}.
-In other words: the high-level model is incomplete from the roles' viewpoint and thus partially valid only.
+However, when the concept of a role appears in the scene, the high-level model is just \textbf{an abstraction of the information handled by the roles}.
+In other words: the high-level model is incomplete from the roles' viewpoint and thus partially valid.
 The high-level model evokes the viewpoint of the \textit{senior experimenter} (see this \href{https://zenodo.org/record/7102464#.YyxvsezMLUK}{\ul{link}}), but with fewer details.
-In turn, the \textit{experiment manager} (see this \href{https://zenodo.org/record/7102450#.Yyxv6ezMLUL}{\ul{link}}) and the \textit{research manager} (see this \href{https://zenodo.org/record/7102431#.YyxwGezMLUL}{\ul{link}}) pay little attention to the design and execution details because the \textit{senior experimenter} takes care of that.
+In turn, the \textit{experiment manager} (see this \href{https://zenodo.org/record/7102450#.Yyxv6ezMLUL}{\ul{link}}), and the \textit{research manager} (see this \href{https://zenodo.org/record/7102431#.YyxwGezMLUL}{\ul{link}}) pay little attention to the design and execution details because the \textit{senior experimenter} takes care of that.
 Likewise, the \textit{experiment manager} is not affected by the experimental research in its global context, e.g., collaboration with other groups and replication.
@@
 There is, apparently, an underlying dichotomy between theory and practice.
-In theory, formal guides, e.g., replication packages, should describe the process step by step.
+In theory, formal guides, e.g., replication packages, should describe the process.
 However, the practice shows that such mechanisms cannot cope with the existing diversity of experimental protocols.
-In our opinion, the replication process will likely benefit from the specialized, fine-grained knowledge \textbf{at the family level}.
+The replication process will likely benefit from the specialized, fine-grained knowledge \textbf{at the family level}.
 \end{itemize}
@@
 We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character.
-Instead, we aimed to observe how experimenters work, and we found out that they perform tasks other than the ones portrayed in textbooks.
+Instead, we aimed to observe how experimenters work and found out that they perform tasks other than those portrayed in textbooks.
 \textbf{This is our essential contribution}.
 
-We are also aware that models represent \textbf{the point of view of a single experimental research group}.
+We also know that models represent \textbf{the point of view of a single experimental research group}.
 We must extend the inquiry to other experimental research groups, ideally to large portions of the ESE community, and check whether they reveal similar behavior patterns.

### 2023-07-25 tex/introduction.tex
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

@@
 Several reasons prevent replication.
-They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE}.
+They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments}, \cite{Demagalhaes-2015-replications-SE}.
 In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.
@@
 Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments.
-Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE} \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data-generation processes vary.
+Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE}, \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.
 
-This paper aims to find out \textbf{how experimental researchers conduct experiments in practice}.
+This paper aims to determine \textbf{how experimental researchers conduct experiments in practice}.
 We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group.
@@
 These models fit the community's procedures and terminology at a high level.
-Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different sub-areas or families of experiments have about the overall process.
+Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.
 
@@
 Section \ref{sec-threats} details the threats to validity and the related mitigation strategies.
-Finally, in section \ref{sec-conclusions}, we present the conclusions and possible future work.
+Finally, in Section \ref{sec-conclusions}, we present the conclusions and future work.

### 2023-07-25 tex/research-execution.tex
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

@@
 \section{Ethnographic study}\label{sec-reseach-execution}
-We report the results of the ethnographic study in chronological order: First, we describe the analysis of the RGUS's preferred literature sources and the review of the RGUS' specific experimental material.
+We report the results of the ethnographic study chronologically: First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material.
 Later, we comment on our experiences replicating an existing experiment.
@@
 \subsection{First Approximation}
-In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support} \cite{Strandberg-2019-Ethical-Interviews-SE} \cite{Yang-2021-interview-study-developers}.
-Accordingly, the researchers started conducting interviews to clarify the RGUS' knowledge and convert it directly into models.
+In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}.
+Accordingly, the researchers conducted interviews to clarify the RGUS' knowledge and convert it directly into models.
 
@@
 The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}.
-We did study the literature.
+Hence, we did study the literature too.
 
 However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments.
-For example, the experimenters supported their story-telling in specific experimental materials that the researchers did not have.
-Nor did the researchers have the knowledge required to understand the experimenters' discussions.
+For example, the experimenters supported their storytelling using specific experimental materials the researchers did not have.
+Nor did the researchers know to understand the experimenters' discussions.
 At this point, the need for a different research methodology --ethnography-- became apparent.
@@
 The interaction with the experimenters also changed.
-For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS families of experiments.
+For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS' families of experiments.
 
@@
 The experimenters use the books by Wohlin et al.
-\cite{Wohlin-2000-Experimentation-SE}, and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process.
+\cite{Wohlin-2000-Experimentation-SE} and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process.
 They also utilize specific papers that advise how to experiment in SE, e.g., Basili et al.
@@
 
-We reviewed the literature in-depth.
+We reviewed the literature in depth.
 We noticed that the sources do not use coherent terminology.
@@
 	\item Wohlin et al.
-\cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report".
-	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution and Analysis".
-	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation".
+\cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report."
+	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution, and Analysis."
+	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation."
 	\item Kitchenham et al.
-\cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data collection, Analysis, Presentation of results, and Interpretation of results".
+\cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data Collection, Analysis, Presentation of results, and Interpretation of results."
 \end{itemize}
@@
 \cite{Kitchenham-2015-Evidence-Based-SE}.
-These books share standard terms, as expected, but they use diverse terms throughout the text, contributing to the terminological diversity.
+As expected, these books share standard terms but use diverse terms throughout the text, contributing to the terminological diversity.
 
@@
 However, accessing the experimental materials was challenging; materials were scattered and disorganized.
-We needed the experimenters' help to find specific documents/data files, understand their structure and contents, and gain further access to other related materials.
+We needed the experimenters' help to find specific documents/data files, understand their structure and contents, and gain further access to related materials.
 
@@
 Experiments vary widely in documentary support, independently of their date.
-The experimental family explains the differences better than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.
+The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.
 
-Concerning the experimental process, all experimental materials describe the activities carried out in a general way.
+Concerning the experimental process, all experimental materials describe the activities carried out generally.
 The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge.
@@
 \subsubsection{Experiment selection}
-Instead of designing an experiment from scratch, the researchers opted to replicate an experiment previously conducted by the RGUS. The replication has been published elsewhere.
+Instead of designing an experiment from scratch, the researchers replicated an experiment previously conducted by the RGUS. The replication has been published elsewhere.
 
 Conducting a replication of the experimenters' previous work attracted their attention and guaranteed collaboration.
-Furthermore, the replication would give us the chance to recreate the design, execution, analysis, and reporting of an experiment that, until then, we only knew due to the examination of the written materials.
+Furthermore, the replication would allow us to recreate the design, execution, analysis, and reporting of an experiment that, until then, we only knew due to the examination of the written materials.
 Finally, the replication would help us better understand the experimenters' motivations and decisions.
@@
 \subsubsection{Replication planning}
-Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE} \cite{Vegas-2006-communication-researchers} \cite{Miller-2005-replicating-SE-experiments} \cite{Gomez-2014-understanding-replication} \cite{Demagalhaes-2015-replications-SE} \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities.
+Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE}, \cite{Vegas-2006-communication-researchers}, \cite{Miller-2005-replicating-SE-experiments}, \cite{Gomez-2014-understanding-replication}, \cite{Demagalhaes-2015-replications-SE}, \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities.
 For example, how to approach the experimental design, which experimental objects are necessary, and in which order.
@@
 The measurement was the most complex and controversial aspect.
-Let's say that the experimental subjects return test cases, but these test cases are manually interpreted and not run.
+Let us say that the experimental subjects return test cases, but these test cases are manually interpreted and not run.
 The experimenters' cooperation was again needed to complete the measurement.
 Even if the measurement procedure looked controversial, it made complete sense after discussion.
-We learned that the experimental protocols depend on the experimental task, which in turn depends on the research area.
+We learned that the experimental protocols depend on the experimental task, which depends on the research area.
 The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty.
@@
 On the other hand, there were reporting guidelines available \cite{Carver-2010-guidelines-replication-SE}.
-The experimenters confirmed that, generally, the analysis and reporting of experiments do not imply substantial SE challenges.
+The experimenters confirmed that the analysis and reporting of experiments generally do not imply substantial SE challenges.
 
 The differences observed between replication and the baseline experiment confirmed our belief that, at least in the case of the RGUS, there are no clear guidelines that allow literal, external, and independent replication \cite{Gomez-2014-understanding-replication} of the original experiment, irrespective of the existence of incidents in the replication's execution.
-As a corollary, the original experimenters' presence was necessary to make the baseline experiment results and the replication comparable.
+As a corollary, the original experimenters' presence was necessary to compare the baseline experiment results and replication.
 
@@
 \subsubsection{Semi-structured interviews}
-We decided to organize new series of semi-structured interviews with the experimenters.
-The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on the replication of \textit{specific families of experiments}.
+We decided to organize a new series of semistructured interviews with the experimenters.
+The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on replicating \textit{specific families of experiments}.
 Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.
@@
 We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences.
-We gave up note-taking and captured the interviews on video, including the experimenters' body language.
+We gave up note-taking and captured the interviews with video, including the experimenters' body language.
 We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative.
@@
 	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
-	\caption{Example of a sketch made by experimenters.}
+	\caption{example of a sketch made by experimenters.}
 	\label{fig-proceso-exp-boseto}
@@
 \begin{itemize}
-	\item \textit{Preliminary} experimental process model: We realized that the way the RGUS conducts experiments resembles the processes we are used to in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}.
+	\item \textit{Preliminary} experimental process model: We realized that the RGUS conducts experiments that resemble the processes we use in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}.
 In the end, they all are processes.
@@
 Each process has several related activities.
-For example, the \textit{Basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).
+For example, the \textit{basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).
 
@@
 In turn, different responsibilities are typically assigned to the same experimenter.
-We identified the following roles (the names suggest their responsibilities): (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}.
-These roles are responsible for: (1) planning the experimental research, (2) managing the logistics, and (3) conducting the experiment in practice, respectively.
+We identified the following roles (the names suggest their responsibilities): (1) \textit{Research Manager}, (2) \textit{Experiment Manager}, and (3) \textit{Senior Experimenter}.
+These roles are responsible for (1) planning the experimental research, (2) managing the logistics, and (3) conducting the experiment in practice, respectively.
 The study showed overlapping responsibilities among roles, e.g., the domain and topic selection for the experiment.
@@
 
-As we said above, we organized the semi-structured interviews around families of experiments.
+As we said above, we organized the semistructured interviews around families of experiments.
 It was a fortunate mistake.
 We found out that the families of experiments are pretty homogeneous.
-The experimenters involved in a family quickly agree and provide a coherent picture of the experimental process.
+The experimenters involved in the family quickly agree and provide a coherent picture of the experimental process.
 However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families.
@@
 Each session lasted at least two hours.
-Before the focus groups, the experimenters reviewed the concept models.
+Before the focus group, the experimenters reviewed the concept models.
 We started with the concept models because they were more straightforward and smaller.
@@
 
-The focus groups prompted a substantial growth in the models' details after heated discussions that re-confirmed the existence of terminological diversity \textit{even among experimenters who carry out similar activities within the same family of experiments}.
+The focus groups prompted a substantial growth in the models' details after heated discussions that reconfirmed the existence of terminological diversity \textit{even among experimenters who carried out similar activities within the same family of experiments}.
 The most likely reason seems to be the training that each experimenter received.
 Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions.
-The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches} \cite{montgomery-2019-Design-Analysis-Experiments} did not drive self-education.
+The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches}, \cite{montgomery-2019-Design-Analysis-Experiments}, did not drive self-education.
 Few of these materials existed when the experimenters started their careers.
@@
 The experimenters did not reach a consensus during the focus group exercise.
-The discussions with experimenters of different families exhibited a characteristic circular pattern: An experimenter triggered a change that, later, was undone by another.
+The discussions with experimenters of different families exhibited a characteristic circular pattern: An experimenter triggered a change later undone by another.
 Substantial agreement was only possible \textit{within each family of experiments}.
@@
 Albeit enjoyable, we perceived it as future research.
-In turn, we tried to take advantage of the limited PhD time available to improve the models by switching the perspective from experimental families to \textit{experimental roles}.
+We tried to use the limited Ph.D.
+time to improve the models by switching the perspective from experimental family to \textit{experimental role}.
 There was no reason to think that roles could contribute to the agreement.
 However, we used the concept model to drive the focus groups, not the process models.
-Roles are connected with activities, not concepts.
+Roles are connected to activities, not concepts.
 Activities looked more uniform than concepts.
@@
 We could also break down the conceptual model into role-specific conceptual models.
-They represent groups of concepts used by the experimenters when they play a specific role.
+They represent groups of concepts the experimenters use when they play a specific role.
 Most experimental activities are role-specific, e.g., the senior experimenter role uses the run-kit concept exclusively.

### 2023-07-25 tex/research-method.tex
Last changes
These are the last changes previous the submission

@@
 
-Ethnography identifies details that are ignored with other methods.
+Ethnography identifies details that are ignored by or in other methods.
 The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}.
@@
 This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research.
-We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.
+We selected a representative i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.
 
-We followed Runenson et al.'s case studies guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:
+We followed Runenson et al.'s case study guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:
 
@@
 To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters".
-We will talk about ourselves the authors as the "the researchers".
-We the authors/researchers play two different roles in the investigation:
+We will talk about ourselves the authors as "the researchers".
+We authors/researchers play two different roles in the investigation:
 
@@
 Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D.
-This is the reason why he is not authoring this paper.
+This is the reason he is not authoring this paper.
 The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R.
@@
 The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information.
-Such taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}.
-Table \ref{tbl-tecnica-fuente} describes the information-gathering techniques and instruments and the corresponding information sources to which they can be applied.
+Such a taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}.
+Table \ref{tbl-tecnica-fuente} describes the information gathering techniques and instruments and the corresponding information sources to which they can be applied.
 
@@
 The order in which researchers contact information sources is essentially random.
-The frequency of contact depends on the researchers' needs to answer research questions during the ethnographic research and the availability of each source.
+The frequency of contact depends on the researchers' need to answer research questions during the ethnographic research and the availability of each source.
 
@@
 \begin{itemize}
-\item \textbf{Preparation of materials}: The information sources are classified and categorized so experimenters and researchers can manage them.
+\item \textbf{Preparation of materials}: The information sources are classified and categorized, so experimenters and researchers can manage them.
 \item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
@@
 \item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters.
-The expected results will be models (conceptual and process) and the practices derived from the interviews with the experimenters.
+The expected results will be models (conceptual and process) and practices derived from the interviews with the experimenters.
 \end{itemize}

### 2023-07-25 tex/results.tex
Last changes
These are the last changes previous the submission

@@
 We have not used standard notations, e.g., UML or BPMN because experimental researchers (and probably readers) have different ages, backgrounds, and careers.
-We have valued communication above model formality.
+We have valued the communication above model formality.
 
@@
 In addition, based on the existence (or not) of new moderator variables and new knowledge generated in the last round, there are activities of analysis associated with planning new cycles of experimentation, replication, or synthesis and reporting related to the publication of results, respectively (highlighted in blue).
-Such us processes appear associated with the experimentation and synthesis paths.
+Such processes appear associated with the experimentation and synthesis paths.
 
@@
 Experimentation is the heart of experimental research since replication and synthesis stem from it.
-However, in some circumstances, experimentation may require the previous execution of synthesis (e.g., when choosing the topic of the experiment) or replication (e.g., in adaptation to the context) activities.
+However, in some circumstances, experimentation may require the previous execution of the synthesis (e.g., when choosing the topic of the experiment) or replication (e.g., in adaptation to the context) activities.
 According to the experimenters, the experimentation activities are:
@@
 	\item Creating the experimental kit.
-	\item Executing and post-execution of the experiment.
+	\item Executing and postexecution of the experiment.
 \end{itemize}
@@
 The replication path consists of the following activities:
+
 \begin{itemize}
@@
 \subsubsection{Synthesis path} 
-According to the RGUS experimenters, the synthesis of results can start from systematic literature reviews, systematic mapping studies, or experiments sharing a common goal (a family of experiments).
-In each case, the experimenters apply particular synthesis techniques (e.g., aggregated or individual patient meta-analysis), followed by interpreting the results.
+According to the RGUS' experimenters, results can be synthesized from systematic literature reviews, systematic mapping studies, or experiments sharing a common goal (a family of experiments).
+In each case, the experimenters apply particular synthesis techniques (e.g., aggregated or individual patient meta-analysis) and interpret the results.
 During the interpretation, the researchers can identify moderator variables or generalize pieces of knowledge.
-Such findings may lead to a reformulation of the synthesis' goal or the conduction of a new cycle of experimental research.
+Such findings may lead to reforming the synthesis' goals or conducting a new cycle of experimental research.
 In addition, the experimenters generally publish results after a synthesis, hand in hand with planning the next round.
@@
 That is correct.
-Our findings are somewhat predictable because the RGUS did not have any experimenter with the required expert knowledge available during this research.
+Our findings are somewhat predictable because the RGUS had no experimenters with the required expert knowledge available during this research.
 On the other hand, the limited insight into synthesis probably represents an indirect confirmation of the ethnographic research's success.
-Our understanding goes in proportion to the expert knowledge available, and there is a lot of specialized knowledge on the RGUS about experimentation and replication, which was our primary goal.
+Our understanding goes in proportion to the expert knowledge available, and there is much-specialized knowledge on the RGUS about experimentation and replication, which was our primary goal.
 
 \subsection{Experimental research concept model}
-This research's conceptual model is not monolithic; on the contrary, it is composed of three sub-models (more accurately, viewpoints) aligned with the prominent experimental roles (research manager - RM, experiment manager - EM, and senior experimenter - SrEx).
+This research's conceptual model is not monolithic; on the contrary, it is composed of three submodels (more accurately, viewpoints) aligned with prominent experimental roles (research manager - RM, experiment manager - EM, and senior experimenter - SrEx).
 The activities described in the EPW lead to such views.
-The experimenters took into account the best-known generic phases of the experimentation process to develop the new versions of the conceptual model (starting from version eight, generated in the first phase of interviews; see \href{https://zenodo.org/record/7102387#.Yyt7W-zMLUI}{\ul{link}}).
+The experimenters took into account the best-known generic phases of the experimentation process to develop new versions of the conceptual model (starting from version eight, generated in the first phase of the interviews; see \href{https://zenodo.org/record/7102387#.Yyt7W-zMLUI}{\ul{link}}).
 
 \subsubsection{Research manager (RM) viewpoint}
-The RM viewpoint is available in this \href{https://zenodo.org/record/7102431#.Yyxi1ezMLUJ}{\ul{link}}. The RM has the following perspective:
+The RM viewpoint is available in this \href{https://zenodo.org/record/7102431#.Yyxi1ezMLUJ}{\ul{link}}. The RM has the following perspectives:
 \begin{itemize}
@@
 	\item During the experiment design phase, the RM takes care of the "human resource management" that could "run," "replicate," or be interested in "replicating" the experiment.
-In addition, the RM receives information regarding the "Expected results," "Execution context," and the "new variables of interest."
+In addition, the RM receives information regarding the "Expected results," "Execution context," and "new variables of interest."
 	\item While the experiment runs, the RM monitors events. However, the data acquisition phase is transparent to her.
-	\item During the data analysis phase, the RM learns the "analysis results" and contributes to the definition of the "findings," "publications," and "pieces of knowledge."
+	\item During the data analysis phase, the RM learns the "analysis results" and contributes to the definition of "findings," "publications," and "pieces of knowledge."
 \end{itemize}
@@
 	\item In the data analysis phase, the EM cooperates with the "re-analysis" in search of new "variables of interest" to the RGUS.
-Additionally, the EM can observe the analysis results to suggest findings that could be useful in the publications.
+Additionally, the EM can observe the analysis results to suggest findings that could be useful in publications.
 \end{itemize}
@@
 	\item In the experiment definition phase, the SrEx observes the "experiment definition," hypotheses' "levels," and "metrics" in the first phase of the experimental investigation.
-The SrEx participates in the definition of the "hypothesis," "factors," and the "response variable."
-	\item In the experimental design phase, the SrEx interacts with a large number of entities; hence, it was necessary to divide such a phase into three sub-phases for this role: \textit{experimental operation design}, \textit{experiment design}, and \textit{artifacts generation}.
+The SrEx participates in the definition of the "hypothesis," "factors," and  "response variable."
+	\item In the experimental design phase, the SrEx interacts with a large number of entities; hence, it is necessary to divide such a phase into three subphases for this role: \textit{experimental operation design}, \textit{experiment design}, and \textit{artifact generation}.
     \begin{itemize}
@@
 	    \item During the \textit{experiment design} proper, the SrEx establishes the execution context's "parameters," the "design type" with its "groups" and assigned "experimental subjects," the "levels" defined and their combination per "session" and "period," "experimental instruments," and the "measurement procedure."
-        \item Finally, in the \textit{artifact generation}, the SrEx fundamentally develops the "run-kit," which includes the construction of the instances of the "guides," "experimental instruments," and "experimental objects." Such "instances" are the tools that the "subjects" need to participate in the "experiment."
+        \item Finally, in the \textit{artifact generation}, the SrEx fundamentally develops the "run-kit," which includes the construction of the instances of "guides," "experimental instruments," and "experimental objects." Such "instances" are the tools that the "subjects" need to participate in the "experiment."
 	\end{itemize}
 	\item While the experiment runs, the SrEx records all the events produced to analyze their impact on the experiment results.
-	\item During data acquisition, the SrEx records the resulting collected items, creates the raw data and obtains measurements.
+	\item During data acquisition, the SrEx records the resulting collected items, creates the raw data, and obtains measurements.
 	\item Finally, in the data analysis phase, the SrEx performs the analysis, obtains the findings, and prepares the publications with the RM and EM's help.
@@
 We have created an initial (generic, i.e., generally applicable) experimental process model (EPM) with the help of the most experienced experimenter at RGUS.
-The model is inspired by the software life cycle \cite{ISO-IEC-IEEE-12207}.
+The software life cycle inspired the model \cite{ISO-IEC-IEEE-12207}.
 We have adapted the concepts used in the ISO/IEC 12207 to the SE experimentation realm.
-Most of the items have been taken from the EPW (see Figure \ref{fig-EPM-construction}); a lesser portion derives from the conceptual models.
+Most items have been taken from the EPW (see Figure \ref{fig-EPM-construction}); a lesser portion derives from the conceptual models.
 The complete process model can be obtained in this \href{https://zenodo.org/record/7105096#.YyxoCOzMLUI}{\ul{link}}.
 
-The EPM contains six groups of processes that define the SE experimental research life cycle: (1) Basic process, (2) Support process, (3) Knowledge item generation process, (4) Publication process, (5) Synthesis process, and (6) Organizational process.
+The EPM contains six groups of processes that define the SE experimental research life cycle: (1) the Basic process, (2) the Support process, (3) the Knowledge item generation process, (4) the Publication process, (5) Synthesis process, and (6) Organizational process.
 
 \subsubsection{Basic process}
-The basic process involve fundamental activities that experimenters perform as RM, SrEx, and EM roles before, during, and after running an experiment.
-Before the experiment, the activities carried out focus on its design, e.g., adapting the experiment to the research context.
+The basic process involves fundamental activities performed by the experimenters as RM, SrEx, and EM roles before, during, and after running an experiment.
+Before the experiment, the activities focused on its design, e.g., adapting the experiment to the research context.
 During the experiment, experimenters pay attention to, e.g., the events that occur before, during, and after the experiment's execution session.
@@
 \subsubsection{Support process}
-The support process includes activities that the EM performs to manage the materials and support those interested in performing replications, as:
+The support process includes activities that the EM performs to manage the materials and support those interested in performing replications, such as:
 
@@
 	\item The management of materials before the experiment's execution (e.g., experiment design document), 
-	\item the materials used during the experiment (e.g., templates for data collection), 
-	\item the materials obtained at the end of the experiment (e.g., filled templates and subjects' code.), and
-	\item the Knowledge transference as support style during replication activities.
+	\item The materials used during the experiment (e.g., templates for data collection), 
+	\item The materials obtained at the end of the experiment (e.g., filled templates and subjects' code.), and
+	\item Knowledge transference as a support style during replication activities.
 \end{itemize}
@@
 \subsubsection{Knowledge items generation process}
-The knowledge items generation process consolidate activities carried out by the RM to manage the knowledge acquired after each experiment, e.g., the knowledge generated in each experimentation phase, the knowledge transferred to professionals, or the knowledge generated during the identification of moderator variables.
+The knowledge items generation process consolidates the activities carried out by the RM to manage the knowledge acquired after each experiment, e.g., the knowledge generated in each experimentation phase, the knowledge transferred to professionals, or the knowledge generated during the identification of moderator variables.
 
@@
 \subsubsection{Synthesis process}
-The EM carries out the synthesis processes. There are different types of synthesis with different evidential levels, e.g., meta-analysis, re-analysis, and informal aggregations.
+The EM carries out the synthesis processes. Different types of synthesis have different evidential levels, e.g., meta-analysis, re-analysis, and informal aggregation.
 
 \subsubsection{Organizational process}
-The organizational process include RM's management activities over the SE experimental research cycle, e.g., human resources and publications management.
+The organizational process includes RM's management activities over the SE experimental research cycle, e.g., human resources and publications management.
 Likewise, the EM also performs management activities on the experiment materials.
-Finally, there are activities related to synthesis management that have not been assigned to any role yet.
+Finally, some activities related to synthesis management have not been assigned to any role yet.
 
@@
 
-Figure \ref{fig-EPM-construction} shows a not very good coincidence between the activities of the EPW and those of the EPM since, as indicated before, each model has its purpose; still, we found several relationships.
+Figure \ref{fig-EPM-construction} shows a not-very-good coincidence between the activities of the EPW and those of the EPM since, as indicated before, each model has its purpose; still, we found several relationships.
 
-Though this model is plausible, it is based on the RGUS' perspective alone.
+This model is plausible but based on the RGUS' perspective alone.
 Further research (in this regard, see Section~\ref{sec-conclusions}) is necessary to ensure that the model represents the experimental SE community's experimental process.

### 2023-07-25 tex/threats-to-validity.tex
Last changes
These are the last changes previous the submission

@@
 71–73]{Runenson-2012-case-study-SE} to design the ethnographic research and describe the threats to validity.
-Runeson et al., following Yin \cite{Yin-2009-case-study}, classify threats in four groups: (1) construct validity, (2) internal validity, (3) external validity, and (4) reliability.
+Runeson et al., following Yin \cite{Yin-2009-case-study}, classify threats into four groups: (1) construct validity, (2) internal validity, (3) external validity, and (4) reliability.
 
@@
 
-The ethnographic method is particularly suitable to oppose this type of threat because knowledge acquisition progress at a slow pace.
-On top of that, we used several techniques during the research to create constructs, e.g., reading, observation, interviews, among others.
+The ethnographic method is particularly suitable to oppose this threat because knowledge acquisition progresses gradually.
+On top of that, we used several techniques during the research to create constructs, e.g., reading, observation, and interviews, among others.
 The different approaches increase the chances of identifying or solving misunderstandings between the researchers and the RGUS.
-We also prevented construct validity threats using reliable information sources, e.g., standard literature and the RGUS' experimental material.
+Using reliable information sources, e.g., standard literature and the RGUS' experimental material, we also prevented construct validity threats.
 
@@
 However, this threat does not apply to this ethnographic study, as we do not aim to identify causal relationships but, at most, correlations between phenomena, e.g., when we claim that the number and diversity of experimental tasks explain the existence of roles.
-Furthermore, most of the findings we provide are descriptive, e.g., the disagreements among experimenters of different experimental families.
+Furthermore, most of the findings we provide are descriptive, e.g., disagreements among experimenters of different experimental families.
 
@@
 As mitigation measures, we have followed the research methodology outlined in Section~\ref{sec-research-method} closely to ensure the reliability of this research.
-Moreover, we have disclosed all intermediate and final results in a repository, available under request, to guarantee transparency in the investigation.
+Moreover, we have disclosed all intermediate and final results in a repository available under request to guarantee transparency in the investigation.

### 2023-10-02 Article.tex
Last requirements
Some changes motivated by the Editor Chief's comments.
Tittle change
Last changes of the title

@@
 
-\title{How does an Experimental Software Engineering Lab Work? An Ethnographic Study}
+\title{An ethnographic study of an empirical software engineering lab}
 
@@
 \section*{Acknowledgment}
-This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
+This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
+This work's preprint has been published \cite{Fonseca-2023-experimentation-proccess-SE-unpublished}.
 

### 2023-12-24 tex/conclusion.tex
Some format and figure changes
Here there are some changes in the entire document.

@@
 Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}.
-Our study has shown that such tacit knowledge is present in SE and manifests relatively superficially, e.g., when researchers from different experimental families discuss their mental models.
+Our study has shown that such tacit knowledge is present in SE and manifests superficially, e.g., when researchers from different experimental families discuss their mental models.
 

### 2023-12-24 tex/research-execution.tex
Some format and figure changes
Here there are some changes in the entire document.

@@
 \section{Ethnographic study}\label{sec-reseach-execution}
-We report the results of the ethnographic study chronologically: First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material.
+We report the results of the ethnographic study chronologically.
+First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material.
 Later, we comment on our experiences replicating an existing experiment.
@@
 In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}.
-Accordingly, the researchers conducted interviews to clarify the RGUS' knowledge and convert it directly into models.
+Accordingly, the researchers conducted interviews to clarify the RGUS knowledge and convert it directly into models.
 
-The interviews with RGUS members provided only public, limited knowledge about experimentation.
+The interviews with RGUS members provided only limited public knowledge about experimentation.
 The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}.
-Hence, we did study the literature too.
+Hence, we studied the literature, too.
 
-However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments.
+However, communication became problematic when the interviews progressed into specialized areas, such as concrete families of experiments.
 For example, the experimenters supported their storytelling using specific experimental materials the researchers did not have.
@@
 The interaction with the experimenters also changed.
-For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS' families of experiments.
+For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS families of experiments.
 
@@
 Experiments vary widely in documentary support, independently of their date.
-The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.
+The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data or automatic tools to measure variables versus manual tools.
 
-Concerning the experimental process, all experimental materials describe the activities carried out generally.
+Concerning the experimental process, all experimental materials generally describe the activities carried out.
 The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge.
@@
 
-The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined in advance all relevant aspects.
+The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined all relevant aspects in advance.
 The measurement was the most complex and controversial aspect.
@@
 However, we were not completely satisfied with the ethnographic study.
-We were able to replicate \textbf{one} experiment in \textbf{one} concrete SE area, but nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}.
+We replicated \textbf{one} experiment in \textbf{one} concrete SE area.
+Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}.
 The presence of tacit knowledge represented a severe threat to validity.
@@
 We decided to organize a new series of semistructured interviews with the experimenters.
-The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on replicating \textit{specific families of experiments}.
+The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}.
+Still, we would focus on replicating \textit{specific families of experiments}.
 Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.
@@
 	\centering
-	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
+	\includegraphics[width=\columnwidth]{images/Process-Approximation-Draft}
 	\caption{example of a sketch made by experimenters.}
@@
 We also created a workflow to represent the relations between activities that appeared during the investigation.
-Please notice that we introduce the outstanding characteristics of the models only because the ethnographic research did not end at this point yet.
+Please note that we introduce the outstanding characteristics of the models only because the ethnographic research has not ended yet.
 The models and the workflow have to be validated by the experimenters.
@@
 	\item \textit{Preliminary} experimental process model: We realized that the RGUS conducts experiments that resemble the processes we use in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}.
-In the end, they all are processes.
+In the end, they are all processes.
 Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization.
@@
 It was a fortunate mistake.
-We found out that the families of experiments are pretty homogeneous.
+The families of experiments are pretty homogeneous.
 The experimenters involved in the family quickly agree and provide a coherent picture of the experimental process.

### 2023-12-24 tex/research-method.tex
Some format and figure changes
Here there are some changes in the entire document.

@@
 The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}.
-Thus, we opted for the study of a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods.
+Thus, we opted to study a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods.
 This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research.
-We selected a representative i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.
+We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.
 
@@
 
-To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters".
-We will talk about ourselves the authors as "the researchers".
-We authors/researchers play two different roles in the investigation:
+To avoid confusion, from this point on, we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters".
+We will talk about ourselves, the authors, as "the researchers".
+We, authors/researchers, play two different roles in the investigation:
 
@@
 Edison G.
-Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D.
+Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of his Ph.D.
 This is the reason he is not authoring this paper.
@@
 \subsection{Context selection}
-We have considered an experimental research group as "representative", i.e., a reliable source of information, when it fulfills the following selection criteria:
+We have considered an experimental research group as "representative," i.e., a reliable source of information when it fulfills the following selection criteria:
 
@@
 	\item The ESE community recognizes the experimenters as specialists in their areas.
-	\item The group has a large number of publications in journals of repute and internal documentation, i.e., the group has a verifiable track record.
+	\item The group has many publications in journals of repute and internal documentation, i.e., the group has a verifiable track record.
 \item One of the main drivers of this research is to explore the connection between replication and lab practices.
@@
 Ethnography information gathering is characterized by being cyclical, iterative, and incremental.
-The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information.
+The data acquisition process was based on a taxonomy of information-gathering techniques, categorized according to the level of contact with the primary source of information.
 Such a taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}.
-Table \ref{tbl-tecnica-fuente} describes the information gathering techniques and instruments and the corresponding information sources to which they can be applied.
+Table \ref{tbl-tecnica-fuente} describes the information-gathering techniques and instruments and the corresponding information sources to which they can be applied.
 
@@
 
-The order in which researchers contact information sources is essentially random.
+The order in which researchers contact information sources is random.
 The frequency of contact depends on the researchers' need to answer research questions during the ethnographic research and the availability of each source.
@@
 
-Ethnographic practice involves storing the raw data and then interpreting the findings \cite{Cooper-2007-Sharing-Data-Ethnographic-Research}.
+The ethnographic practice involves storing the raw data and then interpreting the findings \cite{Cooper-2007-Sharing-Data-Ethnographic-Research}.
 In this research, the data analysis procedure consists of four activities:
@@
 \begin{itemize}
-\item \textbf{Preparation of materials}: The information sources are classified and categorized, so experimenters and researchers can manage them.
+\item \textbf{Preparation of materials}: The information sources are classified and categorized so experimenters and researchers can manage them.
 \item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
@@
 The repository is not public because it contains personal data.
-However, the corresponding author will share the repository with interested researchers upon request.
+However, upon request, the corresponding author will share the repository with interested researchers.
 

### 2024-01-03 Response-to-the-reviewers-comments.md
Cambios necesarios para contestar a la primera revision de IET

# Responses to Reviewer Comments

We appreciate the editor and reviewers for the comments we have received. We have considered all the comments in the current revision. The improvements are indicated in the manuscript using track changes. The final version of the manuscript can be obtained by replacing the line in the "Article.tex" file:

> \usepackage[draft]{changes}

with:

> \usepackage[final]{changes}

Below, we detail the responses to the comments and the changes made:

**Comment 1** - Improved figure number 2

*Response to Comment 1* - Thank you very much for this observation. Indeed, the purpose of this figure needed to be clarified. We have improved the figure by adding the abstraction of the process that researchers obtained from the sketch made by the experimenters. Additionally, we have added the following explanation in the caption:

> Example of a sketch made by the experimenters. The researchers have used drafts like this to progress in defining the experimental process workflow described in Section 4.1.

**Comment 2** - Proper formatting is required throughout the manuscripts.

*Response to Comment 2* - We have carefully reviewed the manuscript.

- We ensured that the LaTeX template corresponds to the latest version, available at https://digital-library.theiet.org/journals/author-guide. The IET logo now appears in the upper right corner by adjusting the template to the newest version.
- We corrected the way the bibliography is presented in the Reference Section. Cited articles now appear correctly in brackets.
- We corrected the lists so that the bullets appear in the correct place.
- We reduced the size of the bullets.

Since these changes affect the overall format, they are not marked in the manuscript. Otherwise, the entire text would appear as modified.

**Comment 3** - Please provide limitations and future scope in the conclusion section.

*Response to Comment 3* - Threats to validity are described in Section 6 "Threats to validity." However, threats to validity do not represent the limitations of the work. We appreciate the reviewer's comment. In Section 7 "Conclusions," we have included the following text:

> However, this work is subject to various limitations. The first and most important is the empirical basis obtained. Although ethnographic research has been extensive over time and conducted using multiple techniques, in the end, we only have data from one research group. Furthermore, this group's experimenters have self-trained in the field, meaning they have not received specific academic training. The observed peculiarities could be habits acquired over time that other, more recent groups with formal training do not possess.
>
> We also cannot rule out errors in perception on the researchers' part as well. As mentioned earlier, we had yet to learn of experimentation at the beginning of ethnographic research, which may have led us to misinterpret some behaviors of the experimenters. On the other hand, perhaps the researchers' "fresh" perspective has allowed us to identify our findings.

Regarding future work, we have modified the single paragraph that appeared on the manuscript:

> We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

replacing it with a more explicit and detailed description of future research plans:

> To overcome the limitations of the present investigation, it is necessary to carry out further studies to generalize the results of the current work. Our plans are as follows:
> - Verify the existence of terminological diversity, one of the first aspects that has emerged in the research and is relatively easy to check. We have already started this work. We surveyed SE experimenters from multiple research groups. In the survey, we asked them about the meaning and usage of experimentation-related terms. Preliminary results confirm the existence of terminological diversity.
>- After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology). Preliminary results indicate the absence of terminological diversity and a greater formalization of the experimental process within specific research areas but not at a general level. This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols (with [51] being a notable exception), which could lead to the need for ad-hoc concepts and processes.
>- Both works will likely be published shortly. In addition, we plan to conduct more studies in the future on this topic soon. We still need to decide on the objects under study and the methodologies used since ethnography is very demanding. A possibility to study the impact of specialization in the sciences is to analyze experimental repositories. This represents an indirect but efficient approach in terms of time and effort. Another alternative is conducting qualitative studies using focus groups composed of SE experimental researchers.

**Comment 4** - Please clearly define the objective of manuscripts in the introduction section.

*Response to Comment 4* - Thank you very much. Indeed, the definition of the research objective was a single and short line. We have re-written the research objective more clearly:

> This paper aims to investigate how experimental researchers conduct experiments in practice. We intend to determine how experimental researchers plan, execute, analyze, and report their investigations and what concepts, protocols, and processes they use.

**Comment 5** - Please add some recent references.

*Response to Comment 5* - We have included the following recent references:

> Traini, L.: 'Exploring performance assurance practices and challenges in agile software development: an ethnographic study', Empirical Software Engineering, 2022, 27, (3), pp. 74.

> Azarova, M., Hazoglou, M., Aronoff.Spencer, E.: 'Just slack it: A study of multidisciplinary teamwork based on ethnography and data from online collaborative software', New Media & Society, 2022, 24, (6), pp. 1435–1458.

>S. Yadav, M. S. I. Sudman, P. Kumar Dubey, R. Vijava Srinivas, R. Srisainath and V. Chithra Devi, "Development of an GA-RBF based Model for Penetration of Electric Vehicles and its Projections," 2023 International Conference on Self Sustainable Artificial Intelligence Systems (ICSSAS), Erode, India, 2023, pp. 1-6, doi: 10.1109/ICSSAS57918.2023.10331883.

### 2024-01-03 Article.tex
Last changes
Some minor changes
Cambios necesarios para contestar a la primera revision de IET

@@
 \usepackage[T1]{fontenc}
-\usepackage{comment}
 \usepackage{framed}
@@
 \urlstyle{same}
-%\usepackage{cite}
+
+\usepackage{natbib}
+\setcitestyle{square}
+
 \usepackage{soul}
@@
 \MakeOuterQuote{"}
+
+\usepackage[draft]{changes}
+
 %%%%%%end MyPackages%%%%
+
+%% Para arreglar las viñetas
+
+\setitemize{label={\raisebox{0pt}{\normalsize\textbullet}}, labelindent=0em,labelsep=1em,leftmargin=*}
+
+%% Hay otros cambios en el .cls
 
@@
 
-\title{An ethnographic study of an empirical software engineering lab}
+\supertitle{Submission Template for IET Research Journal Papers}
+
+\title{Investigation of the activities performed by experimental researchers in a software engineering lab using an ethnographic methodology}
 
@@
 
-\begin{comment}
-
-\section{Insert A head here}
-This demo file is intended to serve as a ``starter file''
-for IMAIAI journal papers produced under \LaTeX\ using
-imaiai.cls v1.5e.
-
-\subsection{Insert B head here}
-Subsection text here.
-
-\subsubsection{Insert C head here}
-Subsubsection text here.
-
-\section{Equations}
-
-Sample equations.
-
-%%% Numbered equation
-\begin{align}\label{1.1}
-\begin{split}
-\frac{\partial u(t,x)}{\partial t} &= Au(t,x) \left(1-\frac{u(t,x)}{K}\right)-B\frac{u(t-\tau,x) w(t,x)}{1+Eu(t-\tau,x)},\\
-\frac{\partial w(t,x)}{\partial t} &=\delta \frac{\partial^2w(t,x)}{\partial x^2}-Cw(t,x)+D\frac{u(t-\tau,x)w(t,x)}{1+Eu(t-\tau,x)},
-\end{split}
-\end{align}
-
-\begin{align}\label{1.2}
-\begin{split}
-\frac{dU}{dt} &=\alpha U(t)(\gamma -U(t))-\frac{U(t-\tau)W(t)}{1+U(t-\tau)},\\
-\frac{dW}{dt} &=-W(t)+\beta\frac{U(t-\tau)W(t)}{1+U(t-\tau)}.
-\end{split}
-\end{align}
-
-%%%% Unnumbered equation
-\[
-\frac{\partial(F_1,F_2)}{\partial(c,\omega)}_{(c_0,\omega_0)} = \left|
-\begin{array}{ll}
-\frac{\partial F_1}{\partial c} &\frac{\partial F_1}{\partial \omega} \\\noalign{\vskip3pt}
-\frac{\partial F_2}{\partial c}&\frac{\partial F_2}{\partial \omega}
-\end{array}\right|_{(c_0,\omega_0)}
-\]
-
-\section{Enunciations}
-%%%% Most of the enunciations like theorem, lemma, corollary, proposition, defintion,
-%%%% condition, example, conjecture etc. are defined in the class file.
-
-%%%% If the author wants to add or modify the enunciation style
-%%%% they can define in the preamble as shown below.
-
-%%%% \newtheoremstyle{theorem}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
-%%%% \theoremstyle{theorem}
-%%%% \newtheorem{theorem}{\sc Theorem}[section]
-
-%%%%\newtheoremstyle{corollary}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
-%%%%\theoremstyle{corollary}
-%%%%\newtheorem{corollary}{\sc Corollary}[section]
-
-%%%%\newtheoremstyle{definition}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
-%%%%\theoremstyle{definition}
-%%%%\newtheorem{definition}[theorem]{\sc Definition}
-%%%%
-%%%%\newtheorem{exercise}[theorem]{Exercise}
-
-\begin{theorem}\label{T0.1}
-Assume that $\alpha>0, \gamma>1, \beta>\frac{\gamma+1}{\gamma-1}$.
-Then there exists a small $\tau_1>0$, such that for $\tau\in
-[0,\tau_1)$, if $c$ crosses $c(\tau)$ from the direction of
-to  a small amplitude periodic traveling wave solution of
-(2.1), and the period of $(\check{u}^p(s),\check{w}^p(s))$ is
-\[
-\check{T}(c)=c\cdot \left[\frac{2\pi}{\omega(\tau)}+O(c-c(\tau))\right].
-\]
-\end{theorem}
-
-\section{Figures \& Tables}
-
-The output for figure is:
-
-\begin{figure}[!h]
-%\centering\includegraphics[width=2.5in]{figurename.eps}
-%%%call your figure name in the place "figurename.eps"
-\caption{Insert figure caption here}
-\label{fig_sim}
-\end{figure}
-
- An example of a double column floating figure using two subfigures.
- (The subfig.sty package must be loaded for this to work.)
- The subfigure \verb+\label+ commands are set within each subfloat command, the
- \verb+\label+ for the overall figure must come after \verb+\caption+.
- \verb+\hfil+ must be used as a separator to get equal spacing.
- The subfigure.sty package works much the same way, except \verb+\subfigure+ is
- used instead of \verb+\subfloat+.
-
-%\begin{figure*}[!h]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{figurename.eps}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{figurename.eps}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-
-\vskip2pc
-
-\noindent The output for table is:
-
-\begin{table}[!h]
-\caption{An Example of a Table}%%%Table caption goes here
-\label{table_example}
-\centering
-\begin{tabular}{|c||c|}%%%The number of columns has to be defined here
-\hline
-One & Two\\ %%%% Table body
-\hline
-Three & Four\\%%%% Table body
-\hline
-\end{tabular}
-\end{table}%%%End of the table
-
-\section{Conclusion}
-The conclusion text goes here.
-\end{comment}
-
 \section*{Acknowledgment}
-This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
+This work was supported by the project \replaced{PGC2018-097265-B-I00}{PID2022-137846NB-I00} funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
 This work's preprint has been published \cite{Fonseca-2023-experimentation-proccess-SE-unpublished}.
@@
 \bibliography{bibliography}
-%
-% once the .bbl file has been generated then place the text in your article.
-
-% To get the numbered reference style the author should use [numbib]
-%as an option in the document class.  For example: \documentclass[numbib]{imaiai}
-\begin{comment}
-
-\begin{thebibliography}{9}
-\bibitem{1}
-Vicsek T., Czir\'{o}k A., Ben-Jacob E., Cohen I.: `Novel type of phase transition in a system of self-driven particles', \textit{Phys. Rev. Lett.}, 1995, \textbf{75}, pp.~1226--1229
-
-\bibitem{2}
-Jadbabaie A., Lin J., Morse A.S.: `Coordination of groups of
-mobile autonomous agents using nearest neighbor rules', \textit{IEEE
-Trans. Autom. Control}, 2003, \textbf{48}, pp.~988--1001
-
-\bibitem{3}
-Olfati-Saber R., Murray R.M.: `Consensus problems in networks
-of agents with switching topology and time-delays', \textit{IEEE Trans.
-Autom. Control}, 2004, \textbf{49}, pp.~1520--1533
-\end{thebibliography}
-\end{comment}
 

### 2024-01-03 tex/conclusion.tex
Cambios necesarios para contestar a la primera revision de IET

@@
 
-We believe that the "lab mythology" is worth studying.
+\added{However, this work is subject to various limitations.
+The first and most important is the empirical basis obtained.
+Although ethnographic research has been extensive over time and conducted using multiple techniques, in the end, we only have data from one research group.
+Furthermore, this group's experimenters have self-trained in the field, meaning they have not received specific academic training.
+The observed peculiarities could be habits acquired over time that other, more recent groups with formal training do not possess.}
+
+\added{We also cannot rule out errors in perception on the researchers' part as well.
+As mentioned earlier, we had yet to learn of experimentation at the beginning of ethnographic research, which may have led us to misinterpret some behaviors of the experimenters.
+On the other hand, perhaps the researchers' "fresh" perspective has allowed us to identify our findings.}
+
+\deleted{We believe that the ''lab mythology'' is worth studying.
 If the differential characteristics of the theoretical vs.
-actual experiment process are specific to the research group, the inquiry will end.
+actual experiment process are specific to the research group, the inquiry will simply end.
+However, it is also possible that widespread patterns of behavior show up.
+Bringing such knowledge to the foreground could contribute to improving experimental practice.}
+
+\deleted{We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore,}
+\added{To overcome the limitations of the present investigation,} it is necessary to carry out further studies to generalize the results of the current work. \added{Our plans are as follows:}
+\begin{itemize}
+
+\item \added{Verify the existence of terminological diversity, one of the first aspects that has emerged in the research and is relatively easy to check.
+We have already started this work.
+We surveyed SE experimenters from multiple research groups.
+In the survey, we asked them about the meaning and usage of experimentation-related terms.
+Preliminary results confirm the existence of terminological diversity.}
+
+\item \added{After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology).
+Preliminary results indicate the absence of terminological diversity and a greater formalization of the experimental process within specific research areas but not at a general level.
+This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols (with \cite{Arcuri-2014-randomized} being a notable exception), which could lead to the need for ad-hoc concepts and processes.}
+
+\item \replaced{Both works will likely be published shortly.
+In addition, we plan}{We plan} to conduct more studies \replaced{in the future}{(probably not ethnographic because they require too much time)} on this topic soon.
+\added{We still need to decide on the objects under study and the methodologies used since ethnography is very demanding.
+A possibility to study the impact of specialization in the sciences is to analyze experimental repositories.
+This represents an indirect but efficient approach in terms of time and effort.
+Another alternative is conducting qualitative studies using focus groups composed of SE experimental researchers.}
+
+\end{itemize}
+
+\replaced{In short, w}{W}e believe that the "lab mythology" is worth studying.
+If the differential characteristics of the theoretical vs.
+actual experiment process \replaced{were}{are} specific to the research group, the inquiry \replaced{would}{will } simply end.
 However, it is also possible that widespread patterns of behavior show up.
 Bringing such knowledge to the foreground could contribute to improving experimental practice.
-
-We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community.
-Therefore, it is necessary to carry out further studies to generalize the results of the present work.
-We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography.
-We have also completed a comparative study of the experimental process between the SE and an established engineering discipline.
-Both works will likely be published shortly.
-In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

### 2024-01-03 tex/introduction.tex
Cambios necesarios para contestar a la primera revision de IET

@@
 
-This paper aims to determine \textbf{how experimental researchers conduct experiments in practice}.
+This paper aims to \replaced{investigate}{determine} \textbf{how experimental researchers conduct experiments in practice}.
+\added{We intend to find out how experimental researchers plan, execute, analyze, and report their investigations and what concepts, protocols, and processes they use.}
+
 We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group.

### 2024-01-03 tex/research-execution.tex
Cambios necesarios para contestar a la primera revision de IET

@@
 	\includegraphics[width=\columnwidth]{images/Process-Approximation-Draft}
-	\caption{example of a sketch made by experimenters.}
+	\caption{Example of a sketch made by the experimenters.
+\added{The researchers have used drafts like this to progress in the definition of the experimental process workflow described in Section~\ref{sec-results:process-workflow}.}}
 	\label{fig-proceso-exp-boseto}
+	\highlight{This new figure replaces the previous Figure 2, which was included}
+	\highlight{in the first version of the manuscript.}
 \end{figure}

### 2024-01-03 tex/research-method.tex
Cambios necesarios para contestar a la primera revision de IET

@@
 \section{Research Method}\label{sec-research-method}
-Ethnography is a \textit{participant observation} method.
+
+Ethnography is a \textit{participant observation} method \added{used with some frequency in SE, e.g., \cite{traini2022exploring, azarova2022just}}.
 It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}.
@@
 
-Ethnography identifies details that are ignored by or in other methods.
+Ethnography identifies details that are ignored by \deleted{or in} other method\replaced{ologies, including Artificial Intelligence (AI) \cite{yadav2023development}}{s}.
 The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}.

### 2024-02-18 revisions/v3/Response-to-the-reviewers-comments-2.md
changes for V3

# Responses to the Reviewers' comments

We appreciate the editor and reviewers for the comments we have received. We have considered all the comments in the current revision. The improvements are indicated in the manuscript using track changes.

The final version of the manuscript can be obtained by replacing the line:

> \usepackage[draft]{changes}

 in the "Article.tex" file with:

> \usepackage[final]{changes}

The manuscript is hosted on the following GitHub repository:

> https://github.com/rodrifonsi/ESE-Process

In some cases, the changes could not be highlighted in the document due to limitations of the *changes* package. However, these changes can be identified by performing a diff between the HEAD and HEAD~1 versions of the .tex files using the following command:

> git diff

Below, we detail the responses to the reviewers' comments and the changes made:

**Comment 1** - Please add recent references.

*Response to Comment 1* - We have included the following recent references:

> Shepperd, M., Ajienka, N., Counsell, S.: ‘The role and value of replication in empirical software engineering results’, Information and Software Technology, 2018, 99, pp. 120–132

> Cockburn, A., Dragicevic, P., Besançon, L., Gutwin, C.: ‘Threats of a replication crisis in empirical computer science’, Communications of the ACM, 2020, 63, (8), pp. 70–79

> The International Organization for Standardization: ‘ISO/IEC 20000-1:2018 Information Technology - Service Management - Part 1: Service Management System Requirements’. 2018. https://www.iso.org/standard/70636.html

>European Commission. ‘Researchcomp: The european competence framework for researchers’. 2022. Accessed: 2024-02-18. https://research-and-innovation.ec.europa.eu/document/download/7da29338-37bf-4d51-b5eb-a1571b84c7ad_en?filename=ec_rtd_research-competence-presentation.pdf

**Comment 2** - Add some future scope conclusion section.

*Response to Comment 2* - Thank you for your suggestion. We have created a Section titled *Implications for practice* containing the following information:

> Our findings suggest specific improvements that could be applied to SE experimentation. The first and most evident is transitioning from an experimental process primarily based on the application of experimental designs and the conduct of statistical analyses to a more professional approach based on explicit processes, activities, and roles, as is customary in many areas, e.g., Information Technology \cite{iso20000}.
>
>It could be argued that current SE experimentation already has a well-defined process and activities, as can be seen, e.g., in the works of Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} or Wohlin \textit{et al.} \cite{Wohlin-2000-Experimentation-SE}. However, as we have argued before, such a process and activities represent only an idealization of the activities performed in practice. We refer to more detailed processes and activities (such as the process models we provide in this work).
>
>Detailed processes and activities only make sense within specific research areas or even families of experiments. For example, Arcuri \& Briand \cite{Arcuri-2014-randomized} have published recommendations for the statistical analysis of randomized algorithms. Arcuri \& Briand do not claim that their recommendations have general applicability. Other areas of SE, such as programming or testing, will need specific recommendations. Although this may seem strange or unusual, a cursory examination of experimental literature in other sciences demonstrates the opposite. For example, in Psychology, there are recommendations on how to measure motivation \cite{toure2014measure} or consider experience \cite{hoffman1998can}. In Medicine, there are models of pain \cite{petersen2002pain}. Other examples could be cited, but all of them agree on highlighting the specificities of a research area instead of hiding details in abstractions that ultimately prove inoperative.
>
>Finally, we must add specific competencies to processes, activities, and roles, such as negotiation skills, asset curation, or human resource management. Again, SE is lagging behind established trends in other areas. To cite a very relevant example, the European Union \cite{ResearchComp} has defined a framework containing seven competency areas, 38 competencies, and 389 learning outcomes for researchers. Other institutions have undertaken similar efforts.

We have also separated in specific sections the research *limitations* and the *future work*.

# Other issues

We have updated the *Acknowledgement* section to comply with the requirements imposed by our funding agency. The text of this section now reads:

> This work was supported by the grant PID2022-137846NB-I00 funded by MCIN/AEI/10.13039/501100011033, and by ''ERDF A way of making Europe'', and the Universidad de las Fuerzas Armadas - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).

### 2024-02-18 Article.tex
changes for V3

@@
 \usepackage[draft]{changes}
+\definechangesauthor[name=version2, color=blue]{v2}
+\definechangesauthor[name=version3, color=red]{v3}
 
@@
 \section*{Acknowledgment}
-This work was supported by the project \replaced{PGC2018-097265-B-I00}{PID2022-137846NB-I00} funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
+This work was supported by the \deleted[id=v3]{project} \replaced[id=v3]{grant PID2022-137846NB-I00 funded by MCIN/AEI/10.13039/501100011033, and by ''ERDF A way of making Europe''}{PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER}, and the Universidad de las Fuerzas Armadas - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII).
 This work's preprint has been published \cite{Fonseca-2023-experimentation-proccess-SE-unpublished}.

### 2024-02-18 tex/conclusion.tex
changes for V3

@@
 
-\added{However, this work is subject to various limitations.
+\subsection{Implications for practice}
+
+\added[id=v3]{Our findings suggest specific improvements that could be applied to SE experimentation.
+The first and most evident is transitioning from an experimental process primarily based on the application of experimental designs and the conduct of statistical analyses to a more professional approach based on explicit processes, activities, and roles, as is customary in many areas, e.g., Information Technology \cite{iso20000}.}
+
+\added[id=v3]{It could be argued that current SE experimentation already has a well-defined process and activities, as can be seen, e.g., in the works of Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} or Wohlin \textit{et al.} \cite{Wohlin-2000-Experimentation-SE}.
+However, as we have argued before, such a process and activities represent only an idealization of the activities performed in practice.
+We refer to more detailed processes and activities (such as the process models we provide in this work).}
+
+\added[id=v3]{Detailed processes and activities only make sense within specific research areas or even families of experiments.
+For example, Arcuri \& Briand \cite{Arcuri-2014-randomized} have published recommendations for the statistical analysis of randomized algorithms.
+Arcuri \& Briand do not claim that their recommendations have general applicability.
+Other areas of SE, such as programming or testing, will need specific recommendations.
+Although this may seem strange or unusual, a cursory examination of experimental literature in other sciences demonstrates the opposite.
+For example, in Psychology, there are recommendations on how to measure motivation \cite{toure2014measure} or consider experience \cite{hoffman1998can}.
+In Medicine, there are models of pain \cite{petersen2002pain}.
+Other examples could be cited, but all of them agree on highlighting the specificities of a research area instead of hiding details in abstractions that ultimately prove inoperative.}
+
+\added[id=v3]{Finally, we must add specific competencies to processes, activities, and roles, such as negotiation skills, asset curation, or human resource management.
+Again, SE is lagging behind established trends in other areas.
+To cite a very relevant example, the European Union \cite{ResearchComp} has defined a framework containing seven competency areas, 38 competencies, and 389 learning outcomes for researchers.
+Other institutions have undertaken similar efforts.}
+
+\subsection{Limitations}
+
+\added[id=v2]{However, this work is subject to various limitations.
 The first and most important is the empirical basis obtained.
@@
 
-\added{We also cannot rule out errors in perception on the researchers' part as well.
+\added[id=v2]{We also cannot rule out errors in perception on the researchers' part as well.
 As mentioned earlier, we had yet to learn of experimentation at the beginning of ethnographic research, which may have led us to misinterpret some behaviors of the experimenters.
@@
 
-\deleted{We believe that the ''lab mythology'' is worth studying.
+\subsection{Future work}
+
+\deleted[id=v2]{We believe that the ''lab mythology'' is worth studying.
 If the differential characteristics of the theoretical vs.
@@
 
-\deleted{We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore,}
-\added{To overcome the limitations of the present investigation,} it is necessary to carry out further studies to generalize the results of the current work. \added{Our plans are as follows:}
+\deleted[id=v2]{We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore,}
+\added[id=v2]{To overcome the limitations of the present investigation,} it is necessary to carry out further studies to generalize the results of the current work.
+\added[id=v2]{Our plans are as follows:}
 \begin{itemize}
 
-\item \added{Verify the existence of terminological diversity, one of the first aspects that has emerged in the research and is relatively easy to check.
+\item \added[id=v2]{Verify the existence of terminological diversity, one of the first aspects that have emerged in the research and is relatively easy to check.
 We have already started this work.
@@
 
-\item \added{After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology).
+\item \added[id=v2]{After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology).
 Preliminary results indicate the absence of terminological diversity and a greater formalization of the experimental process within specific research areas but not at a general level.
-This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols (with \cite{Arcuri-2014-randomized} being a notable exception), which could lead to the need for ad-hoc concepts and processes.}
+This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols which could lead to the need for ad-hoc concepts and processes.}
 
-\item \replaced{Both works will likely be published shortly.
-In addition, we plan}{We plan} to conduct more studies \replaced{in the future}{(probably not ethnographic because they require too much time)} on this topic soon.
-\added{We still need to decide on the objects under study and the methodologies used since ethnography is very demanding.
+\end{itemize}
+
+\replaced[id=v2]{Both works will likely be published shortly.
+In addition, we plan}{We plan} to conduct more studies \replaced[id=v2]{in the future}{(probably not ethnographic because they require too much time)} on this topic soon.
+\added[id=v2]{We still need to decide on the objects under study and the methodologies used since ethnography is very demanding.
 A possibility to study the impact of specialization in the sciences is to analyze experimental repositories.
@@
 Another alternative is conducting qualitative studies using focus groups composed of SE experimental researchers.}
-
-\end{itemize}
-
-\replaced{In short, w}{W}e believe that the "lab mythology" is worth studying.
-If the differential characteristics of the theoretical vs.
-actual experiment process \replaced{were}{are} specific to the research group, the inquiry \replaced{would}{will } simply end.
-However, it is also possible that widespread patterns of behavior show up.
-Bringing such knowledge to the foreground could contribute to improving experimental practice.

### 2024-02-18 tex/discussion.tex
changes for V3

@@
 The concept of~\textquotedblleft role\textquotedblright~applied to experimentation is unusual but not entirely new.
-Mohamed et al.
-\cite{Mohamed-1993-roles-ESE} proposed a framework to conduct ESE experiments.
+Mohamed \textit{et al.} \cite{Mohamed-1993-roles-ESE} proposed a framework to conduct ESE experiments.
 This framework was based on a merge between statistical and SE process concepts, \textbf{including explicit roles}.
@@
 To what extent or how should it be done?
-For example, Ferreira et al.
-\cite{Ferreira-2017-planning-experiments} emphasize methodological consistency over discussing methods and techniques.
+For example, Ferreira \textit{et al.} \cite{Ferreira-2017-planning-experiments} emphasize methodological consistency over discussing methods and techniques.
 

### 2024-02-18 tex/introduction.tex
changes for V3

--- 2023-07-25 tex/introduction.tex
@@
 Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives} \cite{Baker-2016-lid-reproducibility}.
-In Empirical Software Engineering (ESE), the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.
+In Empirical Software Engineering (ESE)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.
 
 Several reasons prevent replication.
-They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments}, \cite{Demagalhaes-2015-replications-SE}.
+They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE} \added[id=v3]{\cite{cockburn2020threats}}.
 In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.
@@
 
-This paper aims to determine \textbf{how experimental researchers conduct experiments in practice}.
+This paper aims to \replaced[id=v2]{investigate}{determine} \textbf{how experimental researchers conduct experiments in practice}.
+\added[id=v2]{We intend to find out how experimental researchers plan, execute, analyze, and report their investigations and what concepts, protocols, and processes they use.}
+
 We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group.

### 2024-02-18 tex/research-execution.tex
changes for V3

@@
 \subsubsection{Analysis of the relevant literature}
-The experimenters use the books by Wohlin et al.
-\cite{Wohlin-2000-Experimentation-SE} and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process.
-They also utilize specific papers that advise how to experiment in SE, e.g., Basili et al.
-\cite{Basili-1986-ESE}, and Kitchenham et al.
-\cite{Kitchenham-2002-empirical-research-guidelines-SE}.
-Basili et al.'s work \cite{Basili-1999-families-experiments} was also considered relevant due to the experimenters' interest in carrying out experimental replications.
+The experimenters use the books by Wohlin \textit{et al.} \cite{Wohlin-2000-Experimentation-SE} and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process.
+They also utilize specific papers that advise how to experiment in SE, e.g., Basili \textit{et al.} \cite{Basili-1986-ESE}, and Kitchenham \textit{et al.} \cite{Kitchenham-2002-empirical-research-guidelines-SE}.
+Basili \textit{et al.}'s work \cite{Basili-1999-families-experiments} was also considered relevant due to the experimenters' interest in carrying out experimental replications.
 
@@
 \begin{itemize}
-	\item Wohlin et al.
-\cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report."
+	\item Wohlin \textit{et al.} \cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report."
 	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution, and Analysis."
-	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation."
-	\item Kitchenham et al.
-\cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data Collection, Analysis, Presentation of results, and Interpretation of results."
+	\item Basili \textit{et al.} \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation."
+	\item Kitchenham \textit{et al.} \cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data Collection, Analysis, Presentation of results, and Interpretation of results."
 \end{itemize}
@@
 We looked for additional ESE books to verify whether terminological diversity was a widespread problem.
-We only found a new edition of Wohlin et al.'s book \cite {Wohlin-2012-experimentatio-SE}, and the specialized books by Runeson et al.
-\cite{Runenson-2012-case-study-SE} and Kitchenham et al.
-\cite{Kitchenham-2015-Evidence-Based-SE}.
+We only found a new edition of Wohlin \textit{et al.}'s book \cite {Wohlin-2012-experimentatio-SE}, and the specialized books by Runeson \textit{et al.} \cite{Runenson-2012-case-study-SE} and Kitchenham \textit{et al.} \cite{Kitchenham-2015-Evidence-Based-SE}.
 As expected, these books share standard terms but use diverse terms throughout the text, contributing to the terminological diversity.
@@
 	\caption{Example of a sketch made by the experimenters.
-\added{The researchers have used drafts like this to progress in the definition of the experimental process workflow described in Section~\ref{sec-results:process-workflow}.}}
+\added[id=v2]{The researchers have used drafts like this to progress in the definition of the experimental process workflow described in Section~\ref{sec-results:process-workflow}.}}
 	\label{fig-proceso-exp-boseto}

### 2024-02-18 tex/research-method.tex
changes for V3

--- 2023-12-24 tex/research-method.tex
@@
 \section{Research Method}\label{sec-research-method}
-Ethnography is a \textit{participant observation} method.
+
+Ethnography is a \textit{participant observation} method \added[id=v2]{used with some frequency in SE, e.g., \cite{traini2022exploring, azarova2022just}}.
 It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}.
@@
 
-Ethnography identifies details that are ignored by or in other methods.
+Ethnography identifies details that are ignored by \deleted[id=v2]{or in} other method\replaced[id=v2]{ologies, including Artificial Intelligence (AI) \cite{yadav2023development}}{s}.
 The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}.
@@
 
-We followed Runenson et al.'s case study guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:
+We followed Runenson \textit{et al.}'s case study guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:
 

### 2024-02-18 tex/threats-to-validity.tex
changes for V3

@@
 Validity is the degree to which the research is conducted transparently and without biases.
-We have followed Runeson et al.'s recommendations \cite[p.
+We have followed Runeson \textit{et al.}'s recommendations \cite[p.
 71–73]{Runenson-2012-case-study-SE} to design the ethnographic research and describe the threats to validity.
-Runeson et al., following Yin \cite{Yin-2009-case-study}, classify threats into four groups: (1) construct validity, (2) internal validity, (3) external validity, and (4) reliability.
+Runeson \textit{et al.}, following Yin \cite{Yin-2009-case-study}, classify threats into four groups: (1) construct validity, (2) internal validity, (3) external validity, and (4) reliability.
 

### 2024-06-22 tex/conclusion.tex
Todos los cambios litos, menos la tabla final

@@
 At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes.
-These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes.
-Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}.
+These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k,Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes.
+Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental,loukides2015beyond}.
 Our study has shown that such tacit knowledge is present in SE and manifests superficially, e.g., when researchers from different experimental families discuss their mental models.

### 2024-06-22 tex/discussion.tex
Todos los cambios litos, menos la tabla final

@@
 Although all empirical researchers know these activities, it is surprising that they are seldom mentioned (if ever) in ESE literature.
-To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE}, \cite{Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}.
+To the best of our knowledge, reference texts, e.g., \cite{Wohlin-2000-Experimentation-SE,Juristo-2001-SE-experimentation}, \textbf{do not address these activities beyond marginal notes}.
 Such richness in the number and diversity of activities raises obvious consequences in experimental education and training.
@@
 Such an inquiry could have an impact on the SE community.
-\item The replication problem is related to knowledge transfer \cite{Shull-2004-Knowledge-sharing-issues-SE}.
+\item The replication problem is related to knowledge transfer \cite{Shull-2004-Knowledge-sharing-issues-SE,collins2022collective}.
 Nevertheless, we wonder what exactly does it mean to transfer knowledge?
@@
 The models (process, workflow, and conceptual) and the ethnographic experience have generated some SE experimental process' differential characteristics or viewpoints, e.g., context adaptation and iterative improvement.
-Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE}, \cite{Sjoberg-2005-survey-experiments-SE}, \cite{Shull-2004-Knowledge-sharing-issues-SE}.
+Previous research pointed out some of these characteristics, e.g., \cite{Mohamed-1993-roles-ESE,Sjoberg-2005-survey-experiments-SE,Shull-2004-Knowledge-sharing-issues-SE}.
 We do not wish to provide a detailed account of these differential characteristics herein; that would require a publication of a different character.

### 2024-06-22 tex/introduction.tex
Todos los cambios litos, menos la tabla final

@@
 \section{Introduction}\label{sec-introduction}
-Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives} \cite{Baker-2016-lid-reproducibility}.
+Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives,Baker-2016-lid-reproducibility}.
 In Empirical Software Engineering (ESE)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.
@@
 Several reasons prevent replication.
-They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments} \cite{Demagalhaes-2015-replications-SE} \added[id=v3]{\cite{cockburn2020threats}}.
+They have discussed elsewhere, e.g., \cite{Miller-2005-replicating-SE-experiments,Demagalhaes-2015-replications-SE,cockburn2020threats,mahmood2018reproducibility,dos2022investigating}.
 In this paper, we focus on a potential cause that, to our knowledge, has not been explored so far: How ESE researchers \textit{really} experiment.
@@
 Anecdotic evidence such as conversations with other researchers, review of articles, and direct observation suggests that researchers follow \textit{in practice} different \textit{experimental processes}, i.e., the different activities that ESE researchers perform to conduct experiments.
-Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE}, \cite{Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.
+Experiment reports look relatively uniform due to the existence of reporting guidelines \cite{Carver-2010-guidelines-replication-SE,Jedlitschka-2008-reporting-experiments-SE}, but the underlying data generation processes vary.
 
@@
 
-We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group.
+We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE,zhang2019ethnographic} within an experienced experimental research group.
 We observed the researchers in their lab for two years.

### 2024-06-22 tex/research-execution.tex
Todos los cambios litos, menos la tabla final

@@
 \subsection{First Approximation}
-In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}.
+In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support,Strandberg-2019-Ethical-Interviews-SE,Yang-2021-interview-study-developers,modi2023using,runeson2020challenges}.
 Accordingly, the researchers conducted interviews to clarify the RGUS knowledge and convert it directly into models.
@@
 \subsubsection{Replication planning}
-Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE}, \cite{Vegas-2006-communication-researchers}, \cite{Miller-2005-replicating-SE-experiments}, \cite{Gomez-2014-understanding-replication}, \cite{Demagalhaes-2015-replications-SE}, \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities.
+Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE,Vegas-2006-communication-researchers,Miller-2005-replicating-SE-experiments,Gomez-2014-understanding-replication,Demagalhaes-2015-replications-SE,Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities.
 For example, how to approach the experimental design, which experimental objects are necessary, and in which order.
@@
 We replicated \textbf{one} experiment in \textbf{one} concrete SE area.
-Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}.
+Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE,Gomez-2014-understanding-replication}.
 The presence of tacit knowledge represented a severe threat to validity.
@@
 We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative.
-We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols}.
+We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols,sowden2020verbal}.
 We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.
@@
 Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions.
-The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches}, \cite{montgomery-2019-Design-Analysis-Experiments}, did not drive self-education.
+The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches,montgomery-2019-Design-Analysis-Experiments}, did not drive self-education.
 Few of these materials existed when the experimenters started their careers.

### 2024-06-22 tex/research-method.tex
Todos los cambios litos, menos la tabla final

@@
 
-Ethnography is a \textit{participant observation} method \added[id=v2]{used with some frequency in SE, e.g., \cite{traini2022exploring, azarova2022just}}.
+Ethnography is a \textit{participant observation} method \added[id=v2]{used with some frequency in SE, e.g., \cite{traini2022exploring,azarova2022just,de2022practices,wang2024ethnographic}}.
 It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}.

### 2024-06-24 tex/appendix.tex

\appendices
\section{Acronyms}

\begin{table}[h]
	\caption{Acronyms used in the manuscript}
	\label{table}
	\setlength{\tabcolsep}{3pt}
	\begin{tabular}{|p{25pt}|p{115pt}|p{25pt}|}
	\hline
	Item & Term & Acronym\\
	\hline
	1 & Empirical Software Engineering & ESE\\
	\hline
	2 & Research Group Under Study & RGUS\\
	\hline
	3 & Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente & GrIMPSoftII\\
	\hline
	4 & Research Manager & RM\\
	\hline
	5 & Experiment Manager & EM\\
	\hline
	6 & Senior Experimenter & SrEx\\
	\hline
	7 & Experimental Process Model & EPM\\
	\hline
	8 & Experimental Process Workflow & EPW\\
	\hline
	\multicolumn{3}{p{251pt}}{These acronyms have been used throughout the manuscript.}\\
	\end{tabular}
	\label{tab1}
\end{table}

### 2024-06-25 tex/appendix.tex
Tercera revisision completada (v4)

@@
 \appendices
+
 \section{Acronyms}
 
+% Please add the following required packages to your document preamble:
+% \usepackage{graphicx}
 \begin{table}[h]
-	\caption{Acronyms used in the manuscript}
-	\label{table}
-	\setlength{\tabcolsep}{3pt}
-	\begin{tabular}{|p{25pt}|p{115pt}|p{25pt}|}
+\small
+\center
+\caption{Acronyms used throughout the manuscript}
+\label{tab:acronyms}
+\begin{tabular}{ll}
 \hline
-	Item & Term & Acronym\\
-	\hline
-	1 & Empirical Software Engineering & ESE\\
-	\hline
-	2 & Research Group Under Study & RGUS\\
-	\hline
-	3 & Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente & GrIMPSoftII\\
-	\hline
-	4 & Research Manager & RM\\
-	\hline
-	5 & Experiment Manager & EM\\
-	\hline
-	6 & Senior Experimenter & SrEx\\
-	\hline
-	7 & Experimental Process Model & EPM\\
-	\hline
-	8 & Experimental Process Workflow & EPW\\
-	\hline
-	\multicolumn{3}{p{251pt}}{These acronyms have been used throughout the manuscript.}\\
+\multicolumn{1}{c}{Acronym} & \multicolumn{1}{c}{Term}       \\ \hline
+\added[id=v4]{EM}                          & \added[id=v4]{Experiment Manager}             \\ 
+\added[id=v4]{EPM}                         & \added[id=v4]{Experimental Process Model}     \\ 
+\added[id=v4]{EPW}                         & \added[id=v4]{Experimental Process Workflow}  \\ 
+\added[id=v4]{ESE}                         & \added[id=v4]{Empirical Software Engineering} \\ 
+\added[id=v4]{RGUS}                        & \added[id=v4]{Research Group Under Study}     \\ 
+\added[id=v4]{RM}                          & \added[id=v4]{Research Manager}               \\ 
+\added[id=v4]{SE}                          & \added[id=v4]{Software Engineering}           \\ 
+\added[id=v4]{SrEx}                        & \added[id=v4]{Senior Experimenter}            \\ \hline 
 \end{tabular}
-	\label{tab1}
 \end{table}

### 2024-06-25 tex/introduction.tex
Tercera revisision completada (v4)

@@
 Replication is troublesome in general \cite{Klein-2018-many}, particularly in the social and life sciences \cite{Pashler-2012-perspectives,Baker-2016-lid-reproducibility}.
-In Empirical Software Engineering (ESE)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.
+In Empirical Software Engineering (ESE%
+\footnote{\added[id=v4]{The acronyms used throughout the paper are listed in Table~\ref{tab:acronyms} in the appendix.}}%
+)\added[id=v3]{, there is a consensus that replications are important \cite{shepperd2018role}, but} the number of replications is small \cite{Bezerra-2015-Replication-SE-U-SMS} and the rate of confirmation of previous results is limited \cite{Jorgensen-2016-Incorrects-Results-SEE}.
 

### 2024-06-25 tex/research-method.tex
Tercera revisision completada (v4)

@@
 
-Ethnography identifies details that are ignored by \deleted[id=v2]{or in} other method\replaced[id=v2]{ologies, including Artificial Intelligence (AI) \cite{yadav2023development}}{s}.
+Ethnography identifies details that are ignored by \deleted[id=v2]{or in} other method\replaced[id=v2]{ologies, including Artificial Intelligence \cite{yadav2023development}}{s}.
 The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}.
@@
 The raw data obtained in this study is organized as shown in Figure \ref{fig-info-study}.
-The tree represents a repository of around 20GB of information.
+The tree represents a repository of around 20 GB of information.
 The repository contains pictures, recordings, models, reports, and transcriptions, most in Spanish (the operational language used during the research).
